---
title: 'Initial population size and baseline survival'
author: ""
date: \today
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(mgcv)
options("scipen" = 100, "digits" = 4)
```


# Preamble

This is an Electronic Supplement to the manuscript Schwacke, L.A. et al.  "Modeling population impacts of the Deepwater Horizon oil spill on a long-lived species: implications and recommendations for future environmental disasters"

The master file containing links to all supplementary files related to this paper is [BND_Electronic Supplements master file](BND_ElectronicSupplements.html).

If you make use of any of this material in your work, it would be appreciated if you would [contact Tiago
Marques](mailto:tiago.marques@st-andrews.ac.uk) to let him know.

# Introduction

This document briefly describes the inital population size and the the post oil population averaged survival. 

It simply reads in and then displays the distributions that we use as data.

For how these distributions are generated please see the dedicated spatial capture recapture (SCR) study by Glennie et al. 2021. 

# Population size

```{r}
# Initial population size
Nstart <- read.csv(paste0("InOutBySp/Bottlenose_dolphin_BB/N_boot.csv"), header = TRUE)
Nstart <- Nstart[, 2]
Nstart <- as.vector(Nstart)
```

The mean initial population size considering `r length(Nstart)` available samples from the posterior of the SCR model ranged between `r round(min(Nstart),0)` and `r round(max(Nstart),0)`. The mean population size was `r round(mean(Nstart))` and the median `r round(median(Nstart))` with a 95% confidence interval of `r round(quantile(Nstart,0.025))` -- `r  round(quantile(Nstart,0.975))`. We can plot the distribution of the population size estimates (removing estimates above double the mean for better visualization)

```{r}
index <- Nstart < 2 * mean(Nstart)
hist(Nstart[index], main = "Initial population size", xlab = "Abundance")
```

# Post-spill survival

```{r}
SpostOilBB <- read.csv(file = "InOutBySp/Bottlenose_dolphin_BB/PostOilSurv.csv", header = FALSE, col.names = "S")
SpostOilBB <- as.numeric(SpostOilBB$S)
```


The population average survival considering `r length(SpostOilBB)` parametric resamples from the SCR model ranged between `r round(min(SpostOilBB),3)` and `r round(max(SpostOilBB),3)`. The mean survival was `r round(mean(SpostOilBB),3)` and the median `r round(median(SpostOilBB),3)` with a 95% interval of `r round(quantile(SpostOilBB,0.025),3)` -- `r  round(quantile(SpostOilBB,0.975),3)`. We can plot the distribution of the survival post oil spill estimates 
```{r}
hist(SpostOilBB, main = "Post oil survival", xlab = "Survival probability")
```

# References

Glennie, R.; Thomas, L.; Speakman, T.; Garrison, L.; Takeshita, R. & Schwacke, L. 2021. [Estimating spatially-varying density and time-varying demographics with open population spatial capture-recapture: a photo-ID case study on bottlenose dolphins in Barataria Bay, Louisana, USA](https://arxiv.org/abs/2106.09579) arXiv:2106.09579
