---
title: "Creating Figure 4"
author: ""
date: \today
output:
  html_document: default
  pdf_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
#to NOT disply scientific notation
options(scipen=999)
```

# Preamble

***This will be a general preable common to several documents included as Electronic Supplements to CARMMHA's BB BND injury paper. The exact preamble will be finalized at the time of submission to reflect the submission date and final title, structure, authors etc.***

This is an Electronic Supplement to the paper submited to Conservation Letters on the ??-??-???? by L.H. Schwacke, T.A. Marques, C. Booth, C. Smith, F. Gomez, B. Balmer, A. Barratclough, S. De Guise, L. Garrison, K. Colegrove, J. Morey, K. D. Mullin, B. M. Quigley, P. Rosel, T. Rowles, R. Takeshita,  F. Townsend, T. Speakman, R. Wells, E. Zolman, L. Thomas, entitled "Why bottlenose dolphin populations have still not recovered a decade after the Deepwater Horizon oil spill"

The master file with supplementary material is here: [Electronic Supplements master file](ElectronicSupplements.html).

# Creating figure 4

## Load libraries and source required files:

```{r,message=FALSE,warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(readr)
library(gridExtra)
source("Functions/reqfuns.R")         # most functions are here
source("Functions/SilerFuns.R")       # Siler model functions are here
```

## Loading required external information and files

Figure 1 is created based on the population size distribution and survival distribution that are the focus of Glennie et al., as well as survival probabilities that were described in Schwacke et al. 2017. We load those .csv files here:

```{r,message=FALSE,warning=FALSE}

BaselineSurvival <- read_csv("InputFiles/BaselineSurvival.csv", 
    col_names = FALSE)

BaselineSurvival$oil<- "Baseline"
#para histograma
N_boot <- read_csv("InOutBySp/Bottlenose_dolphin_BB/N_boot.csv")

PostOilSurv <- read_csv("InOutBySp/Bottlenose_dolphin_BB/PostOilSurv.csv", 
    col_names = FALSE)

PostOilSurv$oil<- "Post-oil"

survdata<- rbind(PostOilSurv,BaselineSurvival)
```

as well as the object that records the simulations over time from our paper. We load those object here:

```{r}
#note this loads a number of objects, including "simres" and "injury"
load("InOutBySp/Bottlenose_dolphin_BB/Ttrusimres1000.RData")
```

## Data wrangling

Here we manipulate the objects to make them suitable for ggplot:

```{r}
#data<- data.frame(simres)
nyears <- dim(simres)[2]
nsims <- dim(simres)[3]
ylims1 <- as.data.frame(c(min(c(colSums(simres[, , 1:nsims, 2]), colSums(simres[, , 1:nsims, 1])))*0.9, max(c(colSums(simres[, , 1:nsims, 2]), colSums(simres[, , 1:nsims, 1])))*1.1))

firstyearsnooil<-data.frame(colSums(simres[, , 1, 2]))
firstyearsoil<- as.data.frame(colSums(simres[, , 1, 1]))

firstyearsnooil1<-(colSums(simres[, , 1, 2]))
firstyearsoil1<- (colSums(simres[, , 1, 1]))

#no oil
data1<- as.data.frame(1:nrow(firstyearsnooil))
colnames(data1)<-"year"
data1$population<-firstyearsnooil1
data1$nooil<- firstyearsnooil1

#oil

data2<- as.data.frame(1:nrow(firstyearsoil))
colnames(data2)<-"year"
data2$population<-firstyearsoil1
data2$withoil<-firstyearsoil1

data3<-merge(data1,data2,by="year",all=TRUE)

data1<- as.data.frame(1:nrow(firstyearsnooil))
colnames(data1)<-"year"
data1$population<-firstyearsnooil1
data1$nooil<- firstyearsnooil1
data1$ymax<- data3$population.x
data1$ymin<- data3$population.y
data1$oil<-"nospill"



data2<- as.data.frame(1:nrow(firstyearsoil))
colnames(data2)<-"year"
data2$oil<- "spill"
data2$population<-firstyearsoil1
data2$withoil<-firstyearsoil1
data2$ymax<- data3$population.x
data2$ymin<- data3$population.y
data<- merge(data1,data2,by=c("year","oil","population","ymin","ymax"), all=TRUE)

```

```{r}
data$diff<-data$ymax-data$ymin

data$oil<-factor(data$oil, levels=c("nospill", "spill"))
```


```{r, echo=FALSE}

#no oil
 poptraj = matrix(NA, ncol = nyears, nrow = nsims)
  for(i in 1:nsims) {
    lines<- colSums(simres[, , i, 2])
    poptraj[i, ] = colSums(simres[, , i, 2])
  }
  
 
pra<- data.frame(t(poptraj))


data11<-cbind(data1,pra)


data111<-  data11 %>% 
  pivot_longer(cols=X1:X1000 ,
               names_to = "observ", 
               values_to = "predict", 
               values_drop_na = TRUE) %>% 
  arrange(year,population,nooil,ymax,ymin,oil,observ,predict)


  
 #oil
 poptraj = matrix(NA, ncol = nyears, nrow = nsims)
  for(i in 1:nsims) {
    lines<- colSums(simres[, , i, 1])
    poptraj[i, ] = colSums(simres[, , i, 1])  }
 
pra<- data.frame(t(poptraj))


data22<-cbind(data2,pra)

data222<-  data22 %>% 
  pivot_longer(cols=X1:X1000 ,
               names_to = "observ", 
               values_to = "predict", 
               values_drop_na = TRUE) %>% 
  arrange(year,population,withoil,ymax,ymin,oil,observ,predict)

data0<- merge(data111,data222,by=c("year","oil","population","ymin","ymax","observ","predict"), all=TRUE)

data0$diff<-data0$ymax-data0$ymin

data0$color<-with(data0, ifelse(oil =="nospill", "pink2",
                                      ifelse(oil == "spill","lightblue", NA)))
```

## Making the plot

### Creating individual plots

For initial population size

```{r, fig.width=15,fig.height=10}
numberplot<-ggplot(N_boot, aes(x=x))+geom_histogram(aes(y=..count..),bins=17,colour="black",alpha=0.6)+coord_flip()+ geom_vline(xintercept=mean(N_boot$x), linetype="dashed", 
                color = "red", size=1.2)+xlim(1000,4500)+xlab("Population size")+theme_bw()
```

For survival

```{r, fig.width=15,fig.height=10}
baselsurv<-ggplot(survdata, aes(x=X1,y=oil))+geom_boxplot(aes(fill=oil),color="black",alpha=0.6)+xlab("Survival")+coord_flip()+geom_jitter(aes(colour=oil),shape=".")+
  scale_fill_manual(values=c("darkblue","brown3"))+
  scale_color_manual(values=c("darkblue","brown3"))+theme_bw()+ylab("")+theme(legend.position = "none")
```

For population trajectories

```{r, fig.width=15,fig.height=10}
bestplot1<-ggplot(data,aes(x=year,y=population,color=oil))+ 
  geom_ribbon(aes(ymin = ymin,ymax=ymax), fill = "red", alpha = .3,data = . %>% filter(year<= 20))+ 
  geom_ribbon(aes(ymin = ymin,ymax=ymax), fill = "blue", alpha = .3,data = . %>% filter(ymin>2792.644 & year> 19))+
  geom_segment(aes(x=10,xend=10,y=1957.725,yend=3258.388),colour="black")+
  geom_line(size=1.4)+
  theme_bw()+
  geom_vline(xintercept=20)+
  labs(colour="")+
  scale_color_manual(values=c("darkblue","brown3"))+
  ylab("Predicted population size")+
  xlab("Years since spill")+
  ylim(1000,4500)+ 
  annotate("segment", xend=20, x= 34.5, y= 4100, yend= 3200, arrow = arrow(type="closed", length = unit(0.1, "inches")))+ 
  annotate('label', 
           label = 'atop(atop("", bold("LTR")), atop("Number of years for the population to", "recover with 95% baseline trajectory"))', 
           y = 4200, x = 34.5, parse = T)+ 
  annotate('label', 
           label = 'atop(atop("", "Lost"), atop("cetacean", "years"))', 
           y = 1800, x = 30, parse = T)+
  annotate("segment", xend=12.5, x= 29, y= 1900, yend= 2800, arrow = arrow(type="closed", length = unit(0.1, "inches"))) + 
  annotate('label', 
           label = 'atop(atop("", bold("MPD")), atop("Largest proportional", "decrease"))', 
           y = 4300, x = 7, parse = T)+
  annotate("segment", xend=10, x= 7, y= 4050, yend= 3200, arrow = arrow(type="closed", length = unit(0.1, "inches")))+ theme(legend.position = c(0.92, 0.2))+xlim(0,50)
```

All in one, i.e. figure 4 in the manuscript

```{r, fig.width=15,fig.height=10,message=FALSE,warning=FALSE}
layout_matrix <- matrix(c(1, 1, 2, 2, 4, 3, 3, 4), nrow = 2, byrow = TRUE)
grid.arrange(numberplot,baselsurv,bestplot1, layout_matrix = matrix(c(1,3,2,3), nrow = 2)) 
```

The results paragraph in the paper is:

The initial population size was `r round(mean(N_boot$x))` (95% credible intervals of `r round(quantile(N_boot$x,0.025))`;`r round(quantile(N_boot$x,0.975))`) (figure 4, top left panel). The mean of the survival pre oil spill was `r round(mean(BaselineSurvival$X1),3)` (95% credible intervals of `r round(quantile(BaselineSurvival$X1,prob=0.025),3)`;`r round(quantile(BaselineSurvival$X1,prob=0.975),3)`) while the mean of the survival post oil spill was `r round(mean(PostOilSurv$X1),3)` (95% credible intervals of `r round(quantile(PostOilSurv$X1,prob=0.025),3)`;`r round(quantile(PostOilSurv$X1,prob=0.975),3)`), leading to an average reduction in survival post-oil spill of `r round(mean(PostOilSurv$X1/BaselineSurvival$X1),3)` (95% credible intervals of `r round(quantile(PostOilSurv$X1/BaselineSurvival$X1,prob=0.025),3)`;`r round(quantile(PostOilSurv$X1/BaselineSurvival$X1,prob=0.975),3)`). The average population trajectory under the presence of the oil spill and assuming the oil spill never happened are shown in Figure 4 (bottom panel). There were  `r  round(median(injury$LCY))`  median LCY (95% percentile CI `r round(quantile(injury$LCY, probs=0.025),0)`-`r round(quantile(injury$LCY, probs=0.975),0)`), the maximum proportional decrease was `r  round(median(injury$MPD),1)` % (95% percentile CI `r round(quantile(injury$MPD, probs=0.025),1)`-`r round(quantile(injury$MPD, probs=0.975),1)`) and it took a median of `r  round(median(injury$YTR),1)` % (95% percentile CI `r round(quantile(injury$YTR, probs=0.025),1)`-`r round(quantile(injury$YTR, probs=0.975),1)`) years for 95 % recovery.

# Aknowledgments

We would like to thank Carolina Marques for the first draft of code and data wrangling that made this plot.
