---
title: 'BND BB Model evaluation: sensitivity and elasticity analysis'
author: ""
date: \today
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(mgcv)
library(knitr)
```


# Preamble

***This will be a general preable common to several documents included as Electronic Supplements to CARMMHA's BB BND injury paper. The exact preamble will be finalized at the time of submission to reflect the submission date and final title, structure, authors etc.***

This is an Electronic Supplement to the paper submited to Conservation Letters on the ??-??-???? by L.H. Schwacke, T.A. Marques, C. Booth, C. Smith, F. Gomez, B. Balmer, A. Barratclough, S. De Guise, L. Garrison, K. Colegrove, J. Morey, K. D. Mullin, B. M. Quigley, P. Rosel, T. Rowles, R. Takeshita,  F. Townsend, T. Speakman, R. Wells, E. Zolman, L. Thomas, entitled "Why bottlenose dolphin populations have still not recovered a decade after the Deepwater Horizon oil spill"

The master file with supplementary material is here: [BND Electronic Supplements master file](BND_ElectronicSupplements.html).

# Introduction

This working document describes how the injury results reported depend on the different population dynamics model parameters. To do so we considered two types of sensitivity analysis:

* Uncertainty analysis - this is akin to what was done in Schwacke et al. 2017. Each parameter for which a distribution was available was allowed to vary according to the assumed precision while keeping all the other parameters constant at their nominal value (i.e., the mean of the corresponding distributions). This type of analysis allows one to evaluate the relative contribution of the parameter variability in the overall variability in reported statistics.

* Elasticity analysis - this is traditionally an analysis that estimates the proportional change in the population growth rate for a proportional change in a vital rate (i.e. survival, growth or reproduction) (Benton & Grant 1999). Here we extend the concept to evaluate the impact of changing a parameter by a given percentage in a given derived statistic from the model, like the reported injury metrics.

# The model parameters

Here we describe a list of the quantities/parameters/statistics considered in the model and that might therefore have an impact in the injury metrics.

The following are statistics for which we have distributions :

* $N_0$: initial population size - distribution available via parametric bootstrap resampling
* $S_m$ and $S_f$: Age dependent baseline survival for males and females
* $\rho$: density dependence shape parameter - EE based, modelled as a shifted gamma distribution
* $P_{recovery}$: proportion of exposed animals that recover to baseline survival and fecundity levels within their lifetime - EE based, constant across taxonomic units, modelled as a shifted beta distribution
* $AFR$: age at first reproduction - modelled as a gamma distribution
* $F_max$: maximum fecundity - modelled as a beta-PERT distribution
* $F_nom$: nominal fecundity - modelled as a beta-PERT distribution
* $R_{baseline}$: baseline reproductive success - modelled as beta distribution
* $R_1$: post oil spill reproductive success - modelled as beta distribution 
* $S_1$: survival pst oil spill, obtained from a SECR capture recapture study
* $P(marked|age)$: probability of an animal being marked as a function of age

Note that since they result from the same spatial density model over which we overlay the oil footprint, the initial population size and proportion exposed are dependent (see IniPopSizes.html for additional details). That dependency will be broken in the sensitivity analysis.

All these quantities and their respective distributions/parameters are provided in file "SpeciesDefinitionFile4offshore.xlsx".

# Sensitivity Analysis 

## Uncertainty

Consider we have $k$ parameters we want to test sensitivity to. And we have run 10000 iterations to evaluate each of the 3 injury metrics. We have run 500 iterations to evaluate sensitivity. Consider LCY as an example. The mean LCY from the 10000 observations is say $\bar{LCY}$. We represent the $LCYPC^k$ as the proportional change in LCY for parameter $k$. What we have represented in the plot below for parameter $k$, where parameter $k$ is allowed to change but all other parameters are kept at their nominal value (here we considered their mean), are the median and the 2.5 and 97.5 percentiles of the distribution of the following statistics (j=1,2,..,500)

$LCYPC^k_{j}=\frac{LCY^k_{j}-\bar{LCY}}{\bar{LCY}}$

```{r}
#BB BND
#select Tursiops
j<-3
runs4Sens<-500
patt<-paste0(runs4Sens,"Sens")
folders<-list.files(path="InOutBySp")
nfolders<-length(folders)
#print(folders[j])
#get files names
files=list.files(path=paste0("InOutBySp/",folders[j]),pattern = patt)
#check how many files
nfiles<-length(files)
#set up number of plots in figure
par(mfrow=c(1,1),mar=c(4,6,0.5,0.5))
plot(c(0.5,11.5),c(-1,1),type="n",xlab="",main="",xaxt="n",ylab="Proportional change (95% CI)")
abline(h=0,lty=2)
for (i in 1:nfiles){
  #for each parameter
  #load the sensitivity results
  #this loads up a workspace that includes object injury
  # a 3 column (one per injury metric) by nsims iterations data.frame
  load(paste0("InOutBySp/",folders[j],"/",files[i]))
  #get name of parameter in files[i]
  str<-files[i]
  #parameter name start
  startp <- str_locate(string=str, pattern="Sens")[2]+1
  #parameters name end
  endp <- str_locate(string=str, pattern= ".RD")[1]-1
  par.name<-substr(str, startp, endp)
  
  #scale the injury by subtracting mean injury for each injury metric
  sinjury<-scale(injury,scale=FALSE)
  #divide by mean for each injury metric
  sinjury[,1]<-sinjury[,1]/mean(injury[,1],na.rm=TRUE)
  sinjury[,2]<-sinjury[,2]/mean(injury[,2],na.rm=TRUE)
  sinjury[,3]<-sinjury[,3]/mean(injury[,3],na.rm=TRUE)
  #plot results
  #boxplot(sinjury,main=par.name)
  qsinjury<-apply(sinjury,2,quantile,probs=c(0.025,0.5,0.975),na.rm=TRUE)
  segments(i-0.2,qsinjury[1,1],i-0.2,qsinjury[3,1],col=1)
  segments(i,qsinjury[1,2],i,qsinjury[3,2],col=3)
  segments(i+0.2,qsinjury[1,2],i+0.2,qsinjury[3,2],col=4)
  points(c(i-0.2,i,i+0.2),qsinjury[2,],col=c(1,3,4))
  label <- par.name
  if(par.name == "a1r") label <- "AFR"
  if(par.name == "N0") label <- expression(N[0])
  if(par.name == "br") label <- expression(R[baseline])
  if(par.name == "por") label <- expression(R[1])
  if(par.name == "rho") label <- expression(rho)
  if(par.name == "per") label <- expression(P[recovery])
  if(par.name == "spos") label <- expression(S[1])
  if(par.name == "Fmax") label <- expression(F[max])
  if(par.name == "Fnom") label <- expression(F[nom])
  if(par.name == "ascS") label <- expression(S[f,j]+S[m,j])
  if(par.name == "PM") label <- expression(P[marked|age])
  axis(side = 1, at = i,labels = label, las = 2)
  legend("topleft",inset=0.05,legend=names(injury),col=c(1,3,4),lty=1,pch=1)
  }
```


```{r,echo=FALSE,eval=FALSE}
#A deprecated alternative representation where each paramter is visualized in its own plot
#BB BND
#select Tursiops
j<-3
runs4Sens<-500
patt<-paste0(runs4Sens,"Sens")
folders<-list.files(path="InOutBySp")
nfolders<-length(folders)
#print(folders[j])
#get files names
files=list.files(path=paste0("InOutBySp/",folders[j]),pattern = patt)
#check how many files
nfiles<-length(files)
#set up number of plots in figure
par(mfrow=c(3,4),mar=c(4,4,2.5,0.5))
for (i in 1:nfiles){
  #for each parameter
  #load the sensitivity results
  #this loads up a workspace that includes object injury
  # a 3 column (one per injury metric) by nsims iterations data.frame
  load(paste0("InOutBySp/",folders[j],"/",files[i]))
  #get name of parameter in files[i]
  str<-files[i]
  #parameter name start
  startp <- str_locate(string=str, pattern="Sens")[2]+1
  #parameters name end
  endp <- str_locate(string=str, pattern= ".RD")[1]-1
  par.name<-substr(str, startp, endp)
  #scale the injury by subtracting mean injury for each injury metric
  sinjury<-scale(injury,scale=FALSE)
  #divide by mean for each injury metric
  sinjury[,1]<-sinjury[,1]/mean(injury[,1],na.rm=TRUE)
  sinjury[,2]<-sinjury[,2]/mean(injury[,2],na.rm=TRUE)
  sinjury[,3]<-sinjury[,3]/mean(injury[,3],na.rm=TRUE)
  #plot results
  #boxplot(sinjury,main=par.name)
  plot(c(0.5,3.5),range(sinjury,na.rm=TRUE),ylim=c(-1,1),type="n",xlab="",main=par.name,xaxt="n",ylab="Proportional change (95% CI)")
  axis(side=1,at=1:3,labels=names(injury))
  qsinjury<-apply(sinjury,2,quantile,probs=c(0.025,0.5,0.975),na.rm=TRUE)
  for(k in 1:3){
    segments(k,qsinjury[1,k],k,qsinjury[3,k])
  }
  points(c(1,2,3),qsinjury[2,])
  abline(h=0,lty=2)
}
```

## Elasticity

In this analysis we used the values obtained under the uncertainty analysis to model the relative change in the injury metrics as a function of the change in the different input parameters. We considered a generalized additive model (GAM), Gaussian response, and identity link. 

The reported results are virtually insensitive to the specific model chosen (here, a GAM) since the relations are all extremely smooth where it matters to evaluate elasticity, which is in the vicinity of the point estimate (here we considered the mean) of the parameter being evaluated.

The corresponding results are shown below, where we report the percentage change on the 3 injury metrics (Lost Cetacean Years LCY, Years to Recovery YTR and Maximum Proportional Decrease MPD ) when each parameter is varied by 1% around the mean (i.e.  increased or decreased by 0.5%), while all the other parameters are kept at their mean value. That is the number reported on top of each plot for each metric considered.

We first define an object to hold the elasticity measure for each parameter.

```{r}
metrics<-c("LCY","YTR","MPD")
nmetrics <- length(metrics)
elasticity<-data.frame(parID=1:9,parameter=NA,label=NA,LCY=NA,YTR=NA,MPD=NA)
```

Here we calculate the injury metric. This code chunk does not output the results by default, but removing the chunk options `results='hide',fig.show='hide'` the .Rmd will lead the code to show plots that provide insights to what is going on under the hood.

```{r, results='hide', fig.show='hide'}
#set up number of plots in figure
par(mfrow=c(3,3),mar=c(4,4,2.5,0.5))
for (k in 1:nmetrics){
cat(paste0("Elasticity analysis for ",metrics[k]))
#a counter to fill up object "elasticity"
m<-1  
for (i in 1:nfiles){
  #for each parameter
  #load the sensitivity results
  #this loakds up a workspace that includes object injury
  # a 3 column (one per injury metric) by nsims iterations data.frame
  load(paste0("InOutBySp/",folders[j],"/",files[i]))
  #get name of parameter in files[i]
  str<-files[i]
  #parameter name start
  startp <- str_locate(string=str, pattern="Sens")[2]+1
  #parameters name end
  endp <- str_locate(string=str, pattern= ".RD")[1]-1
  par.name<-substr(str, startp, endp)
  #these parameters are vectors - canot evaluate like this
  if(par.name=="ascS") next
  if(par.name=="PM") next
  elasticity$parameter[m] <- par.name
  #if(par.name=="a1r") par.name="AFR"
  #only for Ttru do not evaluate
  if(substr(str, 1, 4)=="Ttru" & par.name=="pe") next
  #plot percent change in metric
  p2spar<-eval(parse(text=paste0("pars2save$",par.name)))
  #select the k_th injury (k = 1,2,3)
  inj<-eval(parse(text=paste0("injury$",metrics[k])))
  label <- par.name
  if(par.name == "a1r") label <- "AFR"
  if(par.name == "N0") label <- expression(N[0])
  if(par.name == "br") label <- expression(R[baseline])
  if(par.name == "por") label <- expression(R[1])
  if(par.name == "rho") label <- expression(rho)
  if(par.name == "per") label <- expression(P[recovery])
  if(par.name == "spos") label <- expression(S[1])
  if(par.name == "Fmax") label <- expression(F[max])
  if(par.name == "Fnom") label <- expression(F[nom])
  if(par.name == "ascS") label <- expression(S[f,j]+S[m,j])
  if(par.name == "PM") label <- expression(P[(marked|age)])
  #collect par name
  elasticity$label[m] <- label
  #plot results
  plot(p2spar, inj, xlab= label,ylab=metrics[k])
  dat4mod <- data.frame(par = p2spar, inj=injury[,k])
  gammod <- gam(inj ~ s(par,k=3), data = dat4mod)
  seqpars<-seq(min(p2spar),max(p2spar),length=500)
  preds<-predict.gam(gammod,newdata=data.frame(par = seqpars),type="response")
  #check model adequate for predictions
  lines(seqpars,preds)
  mpar <- mean(p2spar)
  delta <- c(0.995, 1, 1.005)
  nd <- data.frame(par = mpar*delta)
  preds <- predict(gammod, newdata = nd,type="response")
  m.metrics<-mean(inj)
  100*(preds-m.metrics)/m.metrics
  abline(v=c(mpar*delta),col=c("red","green","red"))
  abline(h=c(preds),col=c("red","green","red"))
  # calculate elasticity meajure for the m_th parameter and k_th injury metric 
  elas <- round(100*(preds[3]-preds[1])/preds[2], 3)
  elasticity[m,3+k] <- elas
  mtext(text= elas, side = 3, cex = 0.8)
  #increase counter
m <- m+1
}
}
```

Finally, looking at the elasticity measures in a plot for all parameters at once

```{r}
par(mfrow=c(1,1),mar=c(4,6,0.5,0.5))
barplot(t(as.matrix(elasticity[,4:6])),col=c(2,3,4),beside=TRUE)
for (i in 1:nrow(elasticity)){
  if(elasticity$parameter[i] == "a1r") label <- "AFR"
  if(elasticity$parameter[i] == "N0") label <- expression(N[0])
  if(elasticity$parameter[i] == "br") label <- expression(R[baseline])
  if(elasticity$parameter[i] == "por") label <- expression(R[1])
  if(elasticity$parameter[i] == "rho") label <- expression(rho)
  if(elasticity$parameter[i] == "per") label <- expression(P[recovery])
  if(elasticity$parameter[i] == "spos") label <- expression(S[1])
  if(elasticity$parameter[i] == "Fmax") label <- expression(F[max])
  if(elasticity$parameter[i] == "Fnom") label <- expression(F[nom])
  if(elasticity$parameter[i] == "ascS") label <- expression(S[f,j]+S[m,j])
  if(elasticity$parameter[i] == "PM") label <- expression(P[marked|age])
  axis(side = 1, at = 2.5+(i-1)*4,labels = label, las = 2)
}
legend("bottomleft",col=2:4,pt.bg=2:4,legend=metrics,inset=0.2,cex=2,pch=22)
```

There are two inputs to the model which are not scalars

* $S_{f,j}$ and $S_{m,j}$, and 
* $P_{(marked|age)}$,

for which the above approach to estimate the elasticity is not possible, since there is no univariate quantity to consider that the injury metric depends on. Considering the average survival and the average $P_{(marked|age)}$ as the corresponding univariate quantities we evaluate the injury metrics against was attempted, but naturally that does not really help, because the same mean value can be obtained with different combinations of the vectors, and different combinations of the vectors will have different impacts on the injury metric. This is visible by removing the options `results='hide',fig.show='hide'` in the chunk below. 

```{r, results='hide', fig.show='hide'}
par(mfrow=c(3,2),mar=c(4,4,2.5,0.5))
for (k in 1:nmetrics){
for (i in 1:nfiles){
  #for each parameter
  #load the sensitivity results
  #this loakds up a workspace that includes object injury
  # a 3 column (one per injury metric) by nsims iterations data.frame
  load(paste0("InOutBySp/",folders[j],"/",files[i]))
  #get name of parameter in files[i]
  str<-files[i]
  #parameter name start
  startp <- str_locate(string=str, pattern="Sens")[2]+1
  #parameters name end
  endp <- str_locate(string=str, pattern= ".RD")[1]-1
  par.name<-substr(str, startp, endp)
  label <- par.name
  if(par.name != "ascS" & par.name != "PM") next
  #plot percent change in metric
  p2spar<-eval(parse(text=paste0("pars2save$",par.name)))
  #select the k_th injury (k = 1,2,3)
  inj<-eval(parse(text=paste0("injury$",metrics[k])))
  if(par.name == "ascS") label <- expression(S[f,j]+S[m,j])
  if(par.name == "PM") label <- expression(P[marked|age])
  # 
  plot(p2spar, inj, xlab= label,ylab=metrics[k])
  dat4mod <- data.frame(par = p2spar, inj=injury[,k])
  gammod <- gam(inj ~ s(par,k=3), data = dat4mod)
  seqpars<-seq(min(p2spar),max(p2spar),length=500)
  preds<-predict.gam(gammod,newdata=data.frame(par = seqpars),type="response")
  #check model adequate for predictions
  lines(seqpars,preds)
  mpar <- mean(p2spar)
  delta <- c(0.995, 1, 1.005)
  nd <- data.frame(par = mpar*delta)
  preds <- predict(gammod, newdata = nd,type="response")
  m.metrics<-mean(inj)
  100*(preds-m.metrics)/m.metrics
  abline(v=c(mpar*delta),col=c("red","green","red"))
  abline(h=c(preds),col=c("red","green","red"))
  mtext(text=round(100*(preds[3]-preds[1])/preds[2],3),side=3,cex=0.8)
}
}
```

Instead, we made some special runs for these two parameters in which we calculated the injury metrics for a small increase and decrease (0.5%) of the values of the entire vector of (1) $S_{f,j}$ and $S_{m,j}$, and (2) $P_{(marked|age)}$.

```{r}
#not implemented yet
```



# References

Benton, T. G. & Grant, A. 1999 Elasticity analysis as an important tool in evolutionary and population ecology *Trends in Ecology & Evolution*  **14**: 467-471

Schwacke, L. H.; Thomas, L.; Wells, R. S.; McFee, W. E.; Hohn, A. A.; Mullin, K. D.; Zolman, E. S.; Quigley, B. M.; Rowles, T. K. & Schwacke, J. H. 2017. Quantifying injury to common bottlenose dolphins from the Deepwater Horizon oil spill using an age-, sex- and class-structured population model *Endangered Species Research* **33**: 265-279




