---
title: 'BND BB Model evaluation: sensitivity and elasticity analysis'
author: ""
date: \today
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(mgcv)
library(knitr)
```


# Preamble

This is an Electronic Supplement to the manuscript Schwacke, L.A. et al.  "Modeling population impacts of the Deepwater Horizon oil spill on a long-lived species: implications and recommendations for future environmental disasters"

The master file containing links to all supplementary files related to this paper is [BND_Electronic Supplements master file](BND_ElectronicSupplements.html).

If you make use of any of this material in your work, it would be appreciated if you would [contact Tiago
Marques](mailto:tiago.marques@st-andrews.ac.uk) to let him know.

# Introduction

This working document describes how the injury results reported depend on the different population dynamics model parameters considered. All the parameters have uncertainty associated with them and it is only natural to ask what would change in the reported values for the injury metrics if the input parameters changed accordingly.

To do so we considered two types of sensitivity analysis:

* Uncertainty analysis - this is akin to what was done in Schwacke et al. 2017. Each parameter for which a distribution was available was allowed to vary according to the assumed precision while keeping all the other parameters constant at their nominal value (i.e., the mean of the corresponding distributions). This type of analysis allows one to evaluate the relative contribution of the parameter variability in the overall variability in reported statistics.

* Elasticity analysis - this is traditionally an analysis that estimates the proportional change in the population growth rate for a proportional change in a vital rate (i.e. survival, growth or reproduction) (Benton & Grant 1999). Here we extend the concept to evaluate the impact of changing a parameter by a given percentage in a given derived statistic from the model, like the reported injury metrics.

These analysis are run by sourcing the file `Run_BND_Ttru_PopSimsSensitivity`.

# The model parameters

Here we provide a complete list of the quantities/parameters/statistics considered in the BND BB population dinamics model, that might therefore have an impact in the injury metrics:

* Derived from the SECR capture recapture study reported by Glennie et al. 2021, with distributions  available via parametric bootstrap resampling; see also [BND_InitialPopulation](BND_InitialPopulation.html) for details.
    - $N_0$: initial population size 
    - $S_1$: post oil spill survival

* Based on the Expert Elicitation exercise, see SI section on Expert Elicitation exercise, for details
    - $\rho$: density dependence shape parameter; modelled as a shifted gamma distribution
    - $P_{recovery}$: proportion of exposed animals that recover to baseline survival and fecundity levels within their lifetime; modelled as a shifted beta distribution

* Based on earlier published work, see SI, Table S2.1 - Input parameters for population model, for additional source details 
    - $S_m$ and $S_f$: a vector of age and sex dependent survival for males and females, based on the Siler model parameters, see [SurvivalReduction](SurvivalReduction.html) for details
    - $AFR$: age at first reproduction; modelled as a gamma distribution; see [AgeOfSexualMaturity](AgeOfSexualMaturity.html) for details
    - $F_max$: maximum fecundity - modelled as a beta-PERT distribution
    - $F_nom$: nominal fecundity - modelled as a beta-PERT distribution
    - $R_{baseline}$: baseline reproductive success - modelled as beta distribution
    - $R_1$: post oil spill reproductive success - modelled as beta distribution 

* Based on unpublished data
    - $P_{marked}$: a vector of probabilities of an animal being marked as a function of age, see [SurvivalReduction](SurvivalReduction.html) for details

All these quantities and their respective distributions/parameters are provided in file `SpeciesDefinitionFile.xlsx`.

# Sensitivity Analysis 

## Uncertainty

Consider we have $k$ parameters we want to test sensitivity to. And we have run 10000 iterations to evaluate each of the 3 injury metrics. We have run 500 iterations to evaluate sensitivity. Consider LCY as an example. The mean LCY from the 10000 observations is say $\bar{LCY}$. We represent the $LCYPC^k$ as the proportional change in LCY for parameter $k$. What we have represented in the plot below for parameter $k$, where parameter $k$ is allowed to change but all other parameters are kept at their nominal value (here we considered their mean), are the median and the 2.5 and 97.5 percentiles of the distribution of the following statistics (j=1,2,..,500)

$LCYPC^k_{j}=\frac{LCY^k_{j}-\bar{LCY}}{\bar{LCY}}$

```{r}
#BB BND
#select Tursiops
j<-3
runs4Sens<-1000
patt<-paste0(runs4Sens,"Sens")
folders<-list.files(path="InOutBySp")
nfolders<-length(folders)
#print(folders[j])
#get files names
files=list.files(path=paste0("InOutBySp/",folders[j]),pattern = patt)
#check how many files
nfiles<-length(files)
#set up number of plots in figure
par(mfrow=c(1,1),mar=c(5,4,0.5,0.5))
plot(c(0.5,11.5),c(-1.1,1.1),type="n",xlab="",main="",xaxt="n",ylab="Proportional change (95% CI)")
abline(h=0,lty=2)
for (i in 1:nfiles){
  #for each parameter
  #load the sensitivity results
  #this loads up a workspace that includes object injury
  # a 3 column (one per injury metric) by nsims iterations data.frame
  load(paste0("InOutBySp/",folders[j],"/",files[i]))
  #get name of parameter in files[i]
  str<-files[i]
  #parameter name start
  startp <- str_locate(string=str, pattern="Sens")[2]+1
  #parameters name end
  endp <- str_locate(string=str, pattern= ".RD")[1]-1
  par.name<-substr(str, startp, endp)
  
  #scale the injury by subtracting mean injury for each injury metric
  sinjury<-scale(injury,scale=FALSE)
  #divide by mean for each injury metric
  sinjury[,1]<-sinjury[,1]/mean(injury[,1],na.rm=TRUE)
  sinjury[,2]<-sinjury[,2]/mean(injury[,2],na.rm=TRUE)
  sinjury[,3]<-sinjury[,3]/mean(injury[,3],na.rm=TRUE)
  #plot results
  #boxplot(sinjury,main=par.name)
  qsinjury<-apply(sinjury,2,quantile,probs=c(0.025,0.5,0.975),na.rm=TRUE)
  segments(i-0.2,qsinjury[1,1],i-0.2,qsinjury[3,1],col=4)
  segments(i,qsinjury[1,2],i,qsinjury[3,2],col=2)
  segments(i+0.2,qsinjury[1,2],i+0.2,qsinjury[3,2],col=1)
  points(c(i-0.2,i,i+0.2),qsinjury[2,],col=c(4,2,1),pch=c(16,15,18))
  label <- par.name
  if(par.name == "a1r") label <- "ASM"
  if(par.name == "N0") label <- expression(N[0])
  if(par.name == "br") label <- expression(R[baseline])
  if(par.name == "por") label <- expression(R[1])
  if(par.name == "rho") label <- expression(rho)
  if(par.name == "per") label <- expression(P[recovery])
  if(par.name == "spos") label <- expression(S[1])
  if(par.name == "Fmax") label <- expression(F[max])
  if(par.name == "Fnom") label <- expression(F[nom])
  if(par.name == "ascS") label <- expression(S[f,j]+S[m,j])
  if(par.name == "PM") label <- expression(P[marked])
  axis(side = 1, at = i,labels = label, las = 2)
  legend("topleft",inset=0.05,legend=names(injury),col=c(4,2,1),lty=1,pch=c(16,15,18))
  }
```


```{r,echo=FALSE,eval=FALSE}
#A deprecated alternative representation where each parameter is visualized in its own plot
#set up number of plots in figure
par(mfrow=c(3,4),mar=c(4,4,2.5,0.5))
for (i in 1:nfiles){
  #for each parameter
  #load the sensitivity results
  #this loads up a workspace that includes object injury
  # a 3 column (one per injury metric) by nsims iterations data.frame
  load(paste0("InOutBySp/",folders[j],"/",files[i]))
  #get name of parameter in files[i]
  str<-files[i]
  #parameter name start
  startp <- str_locate(string=str, pattern="Sens")[2]+1
  #parameters name end
  endp <- str_locate(string=str, pattern= ".RD")[1]-1
  par.name<-substr(str, startp, endp)
  #scale the injury by subtracting mean injury for each injury metric
  sinjury<-scale(injury,scale=FALSE)
  #divide by mean for each injury metric
  sinjury[,1]<-sinjury[,1]/mean(injury[,1],na.rm=TRUE)
  sinjury[,2]<-sinjury[,2]/mean(injury[,2],na.rm=TRUE)
  sinjury[,3]<-sinjury[,3]/mean(injury[,3],na.rm=TRUE)
  #plot results
  #boxplot(sinjury,main=par.name)
  plot(c(0.5,3.5),range(sinjury,na.rm=TRUE),ylim=c(-1,1),type="n",xlab="",main=par.name,xaxt="n",ylab="Proportional change (95% CI)")
  axis(side=1,at=1:3,labels=names(injury))
  qsinjury<-apply(sinjury,2,quantile,probs=c(0.025,0.5,0.975),na.rm=TRUE)
  for(k in 1:3){
    segments(k,qsinjury[1,k],k,qsinjury[3,k])
  }
  points(c(1,2,3),qsinjury[2,])
  abline(h=0,lty=2)
}
```

## Elasticity

In this analysis we used the values obtained under the uncertainty analysis to model the relative change in the injury metrics as a function of the change in the different input parameters. We considered a generalized additive model (GAM), Gaussian response, and identity link. 

The reported results are virtually insensitive to the specific model chosen (here, a GAM) since the relations are all extremely smooth where it matters to evaluate elasticity, which is in the vicinity of the point estimate (here we considered the mean) of the parameter being evaluated.

The corresponding results are shown below, where we report the percentage change on the 3 injury metrics (Lost Cetacean Years LCY, Years to Recovery YTR and Maximum Proportional Decrease MPD ) when each parameter is varied by 1% around the mean (i.e.  increased or decreased by 0.5%), while all the other parameters are kept at their mean value. That is the number reported on top of each plot for each metric considered.

We first define an object to hold the elasticity measure for each parameter.

```{r}
metrics<-c("LCY","YTR","MPD")
nmetrics <- length(metrics)
elasticity<-data.frame(parID=1:11,parameter=NA,label=NA,LCY=NA,YTR=NA,MPD=NA)
```

Here we calculate the injury metric percent change for each change in 1% on each of the 11 parameters. The next code chunk does not output plots by default, but by removing the chunk options `results='hide',fig.show='hide'` the .Rmd will lead the code to show plots that provide insights to what is going on under the hood.

```{r, results='hide', fig.show='hide'}
#set up number of plots in figure
par(mfrow=c(3,3),mar=c(4,4,2.5,0.5))
for (k in 1:nmetrics){
cat(paste0("Elasticity analysis for ",metrics[k]))
#a counter to fill up object "elasticity"
m<-1  
for (i in 1:nfiles){
  #for each parameter
  #load the sensitivity results
  #this loakds up a workspace that includes object injury
  # a 3 column (one per injury metric) by nsims iterations data.frame
  load(paste0("InOutBySp/",folders[j],"/",files[i]))
  #get name of parameter in files[i]
  str<-files[i]
  #parameter name start
  startp <- str_locate(string=str, pattern="Sens")[2]+1
  #parameters name end
  endp <- str_locate(string=str, pattern= ".RD")[1]-1
  par.name<-substr(str, startp, endp)
  elasticity$parameter[m] <- par.name
  #these parameters are vectors - cannot evaluate like this
  if(par.name=="ascS" | par.name=="PM") {
    m<-m+1
    next }
  #if(par.name=="a1r") par.name="AFR"
  #only for Ttru do not evaluate
  if(substr(str, 1, 4)=="Ttru" & par.name=="pe") next
  #plot percent change in metric
  p2spar<-eval(parse(text=paste0("pars2save$",par.name)))
  #select the k_th injury (k = 1,2,3)
  inj<-eval(parse(text=paste0("injury$",metrics[k])))
  label <- par.name
  if(par.name == "a1r") label <- "ASM"
  if(par.name == "N0") label <- expression(N[0])
  if(par.name == "br") label <- expression(R[baseline])
  if(par.name == "por") label <- expression(R[1])
  if(par.name == "rho") label <- expression(rho)
  if(par.name == "per") label <- expression(P[recovery])
  if(par.name == "spos") label <- expression(S[1])
  if(par.name == "Fmax") label <- expression(F[max])
  if(par.name == "Fnom") label <- expression(F[nom])
  if(par.name == "ascS") label <- expression(S[f,j]+S[m,j])
  if(par.name == "PM") label <- expression(P[(marked)])
  #collect par name
  elasticity$label[m] <- label
  #plot results
  plot(p2spar, inj, xlab= label,ylab=metrics[k])
  dat4mod <- data.frame(par = p2spar, inj=injury[,k])
  gammod <- gam(inj ~ s(par,k=3), data = dat4mod)
  seqpars<-seq(min(p2spar),max(p2spar),length=500)
  preds<-predict.gam(gammod,newdata=data.frame(par = seqpars),type="response")
  #check model adequate for predictions
  lines(seqpars,preds)
  mpar <- mean(p2spar)
  delta <- c(0.995, 1, 1.005)
  nd <- data.frame(par = mpar*delta)
  preds <- predict(gammod, newdata = nd,type="response")
  m.metrics<-mean(inj)
  100*(preds-m.metrics)/m.metrics
  abline(v=c(mpar*delta),col=c("red","green","red"))
  abline(h=c(preds),col=c("red","green","red"))
  # calculate elasticity meajure for the m_th parameter and k_th injury metric 
  elas <- round(100*(preds[3]-preds[1])/preds[2], 3)
  elasticity[m,3+k] <- elas
  mtext(text= elas, side = 3, cex = 0.8)
  #increase counter
m <- m+1
}
}
```

Finally, looking at the elasticity measures in a plot for all parameters at once

```{r}
par(mfrow=c(1,1),mar=c(4,6,0.5,0.5))
barplot(t(as.matrix(elasticity[,4:6])),col=c(4,2,1),beside=TRUE,ylab="Percent chage per 1% change in parameter")
for (i in 1:nrow(elasticity)){
  if(elasticity$parameter[i] == "a1r") label <- "ASM"
  if(elasticity$parameter[i] == "N0") label <- expression(N[0])
  if(elasticity$parameter[i] == "br") label <- expression(R[baseline])
  if(elasticity$parameter[i] == "por") label <- expression(R[1])
  if(elasticity$parameter[i] == "rho") label <- expression(rho)
  if(elasticity$parameter[i] == "per") label <- expression(P[recovery])
  if(elasticity$parameter[i] == "spos") label <- expression(S[1])
  if(elasticity$parameter[i] == "Fmax") label <- expression(F[max])
  if(elasticity$parameter[i] == "Fnom") label <- expression(F[nom])
  if(elasticity$parameter[i] == "ascS") label <- expression(S[f,j]+S[m,j])
  if(elasticity$parameter[i] == "PM") label <- expression(P[marked])
  axis(side = 1, at = 2.5+(i-1)*4,labels = label, las = 2)
}
legend("bottomleft",col=c(4,2,1),pt.bg=c(4,2,1),legend=metrics,inset=0.2,cex=1.2,pch=22)
```

Note that in the image above the two inputs to the model which are not scalars, namely

* $S_{f,j}$ and $S_{m,j}$, and 
* $P_{marked}$,

are not shown, since there is no univariate quantity to consider that the injury metric depends on, and hence no obvious value to represent in the x axis. One might be tempted to consider the average survival and the average $P_{marked}$ as the corresponding univariate quantities to evaluate the injury metrics against. However, that does not really help, because the same mean value can be obtained with different combinations of the vectors, and different combinations of the vectors will have different impacts on the injury metric. This is visible by removing the options `results='hide',fig.show='hide'` in the chunk below. 

```{r, results='hide', fig.show='hide'}
par(mfrow=c(3,2),mar=c(4,4,2.5,0.5))
for (k in 1:nmetrics){
for (i in 1:nfiles){
  #for each parameter
  #load the sensitivity results
  #this loakds up a workspace that includes object injury
  # a 3 column (one per injury metric) by nsims iterations data.frame
  load(paste0("InOutBySp/",folders[j],"/",files[i]))
  #get name of parameter in files[i]
  str<-files[i]
  #parameter name start
  startp <- str_locate(string=str, pattern="Sens")[2]+1
  #parameters name end
  endp <- str_locate(string=str, pattern= ".RD")[1]-1
  par.name<-substr(str, startp, endp)
  label <- par.name
  if(par.name != "ascS" & par.name != "PM") next
  #plot percent change in metric
  p2spar<-eval(parse(text=paste0("pars2save$",par.name)))
  #select the k_th injury (k = 1,2,3)
  inj<-eval(parse(text=paste0("injury$",metrics[k])))
  if(par.name == "ascS") label <- expression(S[f,j]+S[m,j])
  if(par.name == "PM") label <- expression(P[marked])
  # 
  plot(p2spar, inj, xlab= label,ylab=metrics[k])
  dat4mod <- data.frame(par = p2spar, inj=injury[,k])
  gammod <- gam(inj ~ s(par,k=3), data = dat4mod)
  seqpars<-seq(min(p2spar),max(p2spar),length=500)
  preds<-predict.gam(gammod,newdata=data.frame(par = seqpars),type="response")
  #check model adequate for predictions
  lines(seqpars,preds)
  mpar <- mean(p2spar)
  delta <- c(0.995, 1, 1.005)
  nd <- data.frame(par = mpar*delta)
  preds <- predict(gammod, newdata = nd,type="response")
  m.metrics<-mean(inj)
  100*(preds-m.metrics)/m.metrics
  abline(v=c(mpar*delta),col=c("red","green","red"))
  abline(h=c(preds),col=c("red","green","red"))
  mtext(text=round(100*(preds[3]-preds[1])/preds[2],3),side=3,cex=0.8)
}
}
```

Instead, we made some special runs of `Run_BND_Ttru_PopSimsSensitivity` for these two parameters in which we calculated the injury metrics for a small increase and decrease (0.5%) of the values of the entire vector of (1) $S_{f,j}$ and $S_{m,j}$, and (2) $P_{marked}$.

```{r}
load("InOutBySp/Bottlenose_dolphin_BB/Ttrusimres3ElasPME.RData")
elasticity[elasticity$parameter=="PM",4:6]<- 100*(injury[3,]-injury[1,])/(injury[2,])
load("InOutBySp/Bottlenose_dolphin_BB/Ttrusimres3ElasascSE.RData")
elasticity[elasticity$parameter=="ascS",4:6]<- 100*(injury[3,]-injury[1,])/(injury[2,])
```

The final plot is here:

```{r}
par(mfrow=c(1,1),mar=c(4,6,0.5,0.5))
barplot(t(as.matrix(elasticity[,4:6])),col=c(4,2,1),beside=TRUE,ylab="Percent chage per 1% change in parameter")
for (i in 1:nrow(elasticity)){
  if(elasticity$parameter[i] == "a1r") label <- "ASM"
  if(elasticity$parameter[i] == "N0") label <- expression(N[0])
  if(elasticity$parameter[i] == "br") label <- expression(R[baseline])
  if(elasticity$parameter[i] == "por") label <- expression(R[1])
  if(elasticity$parameter[i] == "rho") label <- expression(rho)
  if(elasticity$parameter[i] == "per") label <- expression(P[recovery])
  if(elasticity$parameter[i] == "spos") label <- expression(S[1])
  if(elasticity$parameter[i] == "Fmax") label <- expression(F[max])
  if(elasticity$parameter[i] == "Fnom") label <- expression(F[nom])
  if(elasticity$parameter[i] == "ascS") label <- expression(S[f,j]+S[m,j])
  if(elasticity$parameter[i] == "PM") label <- expression(P[marked])
  axis(side = 1, at = 2.5+(i-1)*4,labels = label, las = 2)
}
legend("bottomleft",col=c(4,2,1),pt.bg=c(4,2,1),legend=metrics,inset=0.2,cex=1.2,pch=22)
```


# References

Benton, T. G. & Grant, A. 1999 Elasticity analysis as an important tool in evolutionary and population ecology *Trends in Ecology & Evolution*  **14**: 467-471

Schwacke, L. H.; Thomas, L.; Wells, R. S.; McFee, W. E.; Hohn, A. A.; Mullin, K. D.; Zolman, E. S.; Quigley, B. M.; Rowles, T. K. & Schwacke, J. H. 2017. Quantifying injury to common bottlenose dolphins from the Deepwater Horizon oil spill using an age-, sex- and class-structured population model *Endangered Species Research* **33**: 265-279




