---
title: "Scaling survival: comparing scaling based on gestation duration across species"
author: ""
date: "7/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Preamble

***This will be a general preable common to all the documents included as Electronic Supplements to CARMMHA's offshore injury paper. The exact preamble will be finalized at the time of submission to reflect the submission date and final title, structure, authors etc.***

This is an Electronic Supplement to the paper submited to MEPS on the ??-??-???? by Marques, T. A., Thomas, L., Booth, C., Garrison, L. P., Rosel, P. E., Mullin, K. D. & L. Schwacke entitled "Quantifying Deepwater Horizon oil spill induced injury on offshore cetaceans".

The master file with supplementary material is here: [Electronic Supplements master file](ElectronicSupplements.html).

# Introduction

In this document we look at the age specific survival used for each species once scaled by the gestation duration, as described in the manuscript.

To understand why we ended up here a relevant source is Marques(2020) On scaling survival and the Siler model. A CARMMHA working document. Despite being a working document not prepared for a wider distribution, please contact tiago.marques@st-andrews.ac.uk if you are interested in a copy of said document.

# Reading relevant info

```{r,echo=FALSE}
library(readxl)
library(knitr)
library(ggplot2)
# Siler model functions are here
source("Functions/SilerFuns.R")       
```

We read in the species details

```{r}
#read the species definition file
SDF <- read_excel("InputFiles/SpeciesDefinitionFile4offshore.xlsx",na="NA")
```

and we set the species which we want results for, and then filter the input data for those species only.

```{r}
#remove irrelevant species rows
#SDF <- SDF[-c(1,6,7,14,16,22:27),]
sps<-c("Bwsp","Fatt","Gmac","Ggri","Kosp","Pele","Pmac","Satt","Sbre","Scly","Scoe","Sfro","Slon","Ttro","Ttrs","Ttru") 
SDF <- SDF[SDF$Species %in% sps,]
```

We read in *Tursiops truncatus* bottlenose dolphin (BND) Siler model posterior values. These can then be used to estimate:

1. survival conditional on age, and 
2. survivorship functions for this species. 

The corresponding values for all other species/stocks are obtained by scaling the *Tursiops truncatus* values by gestation duration.

```{r}
#--------------------------------------------------------------------------
# Survival
#--------------------------------------------------------------------------
# reading in Siler parameters, for males and females
# These files were provided by LS via email on the Fri 5/22/2020 4:10 PM
pf <- read.csv("InputFiles/var_female 15May2020.csv")
pm <- read.csv("InputFiles/var_male 15May2020.csv")
# remove useless ID first column 
pf<-pf[, -1]
pm<-pm[, -1]
# rename columns
names(pf) <- names(pm) <- c("a1", "a2", "a3", "b1", "b3", "rmean")
#--------------------------------------------------------------------------
```

# *Tursiops truncatus*

Here we represent the relevant functions and the associated variability is also illustrated.

## Survivorship function

Plotting the age specific survival aka survivorship function for BND

```{r,cache=TRUE,fig.width = 10}
ages <- 0:70
ncols <- nrow(pf)
nrows <- length(ages)
#get a dataframe to use ggplot2
#objects to hold female and male realizations
lxFs <- matrix(NA,nrow=nrows,ncol=ncols)
lxMs <- matrix(NA,nrow=nrows,ncol=ncols)
#for each obseration of the posterior
for(i in 1:nrow(pf)){
  #get the funtion
  lxFs[,i] <- lx(ages,pf[i,1],pf[i,2],pf[i,3],pf[i,4],pf[i,5])
  lxMs[,i] <- lx(ages,pm[i,1],pm[i,2],pm[i,3],pm[i,4],pm[i,5])
}
#arrange as data frame
lxFs2 <- data.frame(ages=ages,lxFs)
lxMs2 <- data.frame(ages=ages,lxMs)
# add means and relevant quantiles
lxFs2$mean=rowMeans(lxFs)
lxMs2$mean=rowMeans(lxMs)
lxFs2$q025=apply(X=lxFs,MARGIN = 1,FUN=quantile,probs=0.025)
lxMs2$q025=apply(X=lxMs,MARGIN = 1,FUN=quantile,probs=0.025)
lxFs2$q975=apply(X=lxFs,MARGIN = 1,FUN=quantile,probs=0.975)
lxMs2$q975=apply(X=lxMs,MARGIN = 1,FUN=quantile,probs=0.975)
#add the sex
lxFs2$sex="Female"
lxMs2$sex="Male"
#make single object
lxMF <- rbind(lxFs2,lxMs2)
#plot
#https://www.r-graph-gallery.com/104-plot-lines-with-error-envelopes-ggplot2.html
ggplot(data=lxMF, aes(x=ages, y=mean, ymin=q025, ymax=q975, fill=sex, linetype=sex)) + geom_line() + 
 geom_ribbon(alpha=0.5) + 
 xlab("Age") + 
 ylab("Proportion surviving")
```

# Age specific Survival

```{r,cache=TRUE,fig.width = 10}
ages <- 0:70
ncols <- nrow(pf)
nrows <- length(ages)
#get a dataframe to use ggplot2
#objects to hold female and male realizations
pxFs <- matrix(NA,nrow=nrows,ncol=ncols)
pxMs <- matrix(NA,nrow=nrows,ncol=ncols)
#for each obseration of the posterior
for(i in 1:nrow(pf)){
  #get the funtion
  pxFs[,i] <- px(ages,pf[i,1],pf[i,2],pf[i,3],pf[i,4],pf[i,5])
  pxMs[,i] <- px(ages,pm[i,1],pm[i,2],pm[i,3],pm[i,4],pm[i,5])
}
#arrange as data frame
pxFs2 <- data.frame(ages=ages,pxFs)
pxMs2 <- data.frame(ages=ages,pxMs)
# add means and relevant quantiles
pxFs2$mean=rowMeans(pxFs)
pxMs2$mean=rowMeans(pxMs)
pxFs2$q025=apply(X=pxFs,MARGIN = 1,FUN=quantile,probs=0.025)
#two out of 4000*71 values are NA's?
pxMs2$q025=apply(X=pxMs,MARGIN = 1,FUN=quantile,probs=0.025,na.rm=TRUE)
pxFs2$q975=apply(X=pxFs,MARGIN = 1,FUN=quantile,probs=0.975)
pxMs2$q975=apply(X=pxMs,MARGIN = 1,FUN=quantile,probs=0.975,na.rm=TRUE)
#add the sex
pxFs2$sex="Female"
pxMs2$sex="Male"
#make single object
pxMF <- rbind(pxFs2,pxMs2)
#plot
#https://www.r-graph-gallery.com/104-plot-lines-with-error-envelopes-ggplot2.html
ggplot(data=pxMF, aes(x=ages, y=mean, ymin=q025, ymax=q975, fill=sex, linetype=sex)) + geom_line() + 
 geom_ribbon(alpha=0.5) + 
 xlab("Age") + 
 ylab("Age specific Survival")
```

# Comparing all species

The scaling is done by gestation duration, so we look at the gestation duration considered for each of the species. The rationale for the gestation duration used is provided on a separate document included as a separate Electronic Supplement (ElectronicSupplement4_GD).

The range in gestation duration is shown below, plotted against the proportion of the population exposed to oil (this second variable was chosen somewhat arbitrarily, just to allow a good spread of the data in a biplot).

```{r, fig.width = 10, fig.height = 6}
pes<-read.table(file="Pe4CSAS.txt",header=TRUE)
plot(pes$x, SDF$gd, ylab = "Gestation duration", xlab = "Proportion exposed", type = "n")
text(pes$x, SDF$gd, labels = SDF$Species)
```

# Survivorship function

We present the scalled survivorship by function below. For reference BB BND shown in green, sperm whale in blue and spinner dolphin in light blue

```{r, fig.width = 10,fig.height=6}
par(mfrow = c(2, 1),mar = c(4, 4, 0.5, 0.5))
ages <- 0:70
# Proportion surviving to age x
plot(ages, lx(ages*1, pf[1, 1], pf[1, 2], pf[1, 3], pf[1, 4], pf[1, 5]), type = "l", main = "", ylab = "Proportion", xlab = "Age (females)", lty = 1)
for(i in 1:nrow(SDF)){
  # Proportion surviving to age x
lines(ages, lx(ages*(SDF$gd[1]/SDF$gd[i]), pf[1, 1], pf[1, 2], pf[1, 3], pf[1, 4], pf[1, 5]), type="l", main="Proportion surviving to age x", ylab="Proportion", xlab="Age (females)", lty=i,col=ifelse(SDF$Species[i]=="Ttru",3,ifelse(SDF$Species[i]=="Pmac",4,ifelse(SDF$Species[i]=="Slon",5,1))),lwd=ifelse(SDF$Species[i]=="Ttru",2,ifelse(SDF$Species[i]=="Pmac",2,ifelse(SDF$Species[i]=="Slon",2,1))))
}
plot(ages, lx(ages*1 , pm[1, 1], pm[1, 2], pm[1, 3], pm[1, 4], pm[1, 5]), type = "l", lty = 1, xlab="Age (males)", ylab="Proportion")
for(i in 1:nrow(SDF)){
# Proportion surviving to age x
lines(ages, lx(ages*(SDF$gd[1]/SDF$gd[i]), pm[1, 1], pm[1, 2], pm[1, 3], pm[1, 4], pm[1, 5]), lty=i, xlab="Age (males)",col=ifelse(SDF$Species[i]=="Ttru",3,ifelse(SDF$Species[i]=="Pmac",4,ifelse(SDF$Species[i]=="Slon",5,1))),lwd=ifelse(SDF$Species[i]=="Ttru",2,ifelse(SDF$Species[i]=="Pmac",2,ifelse(SDF$Species[i]=="Slon",2,1))))
}
legend("topright",legend=c("Pmac","Ttru","Slon"),col=c(4,3,5),lwd=2,inset=0.05)
```

# Age specific survival

We present the scalled age specific survival by function below. For reference BB BND shown in green, sperm whale in blue and spinner dolphin in light blue

```{r, fig.width = 10,fig.height=6}
par(mfrow = c(2, 1),mar = c(4, 4, 0.5, 0.5))
ages <- 0:140
# Proportion surviving to age x
plot(ages, px(ages*1, pf[1, 1], pf[1, 2], pf[1, 3], pf[1, 4], pf[1, 5],scaling=1), type = "l", main = "", ylab = "Survival", xlab = "Age (females)", lty = 1, ylim = c(0, 1))
for(i in 1:nrow(SDF)){
  # Proportion surviving to age x
lines(ages, px(ages*(SDF$gd[1]/SDF$gd[i]), pf[1, 1], pf[1, 2], pf[1, 3], pf[1, 4], pf[1, 5],scaling=(SDF$gd[1]/SDF$gd[i])), type="l", main="Proportion surviving to age x", ylab="Proportion", xlab="Age (females)", lty=i,col=ifelse(SDF$Species[i]=="Ttru",3,ifelse(SDF$Species[i]=="Pmac",4,ifelse(SDF$Species[i]=="Slon",5,1))),lwd=ifelse(SDF$Species[i]=="Ttru",2,ifelse(SDF$Species[i]=="Pmac",2,ifelse(SDF$Species[i]=="Slon",2,1))))
}
plot(ages, px(ages*1 , pm[1, 1], pm[1, 2], pm[1, 3], pm[1, 4], pm[1, 5],scaling=1), type = "l", lty = 1, xlab="Age (males)", ylab="Survival", ylim = c(0, 1))
for(i in 1:nrow(SDF)){
# Proportion surviving to age x
lines(ages, px(ages*(SDF$gd[1]/SDF$gd[i]), pm[1, 1], pm[1, 2], pm[1, 3], pm[1, 4], pm[1, 5],scaling=(SDF$gd[1]/SDF$gd[i])), lty=i, xlab="Age (males)",col=ifelse(SDF$Species[i]=="Ttru",3,ifelse(SDF$Species[i]=="Pmac",4,ifelse(SDF$Species[i]=="Slon",5,1))),lwd=ifelse(SDF$Species[i]=="Ttru",2,ifelse(SDF$Species[i]=="Pmac",2,ifelse(SDF$Species[i]=="Slon",2,1))))
}
legend("topright",legend=c("Pmac","Ttru","Slon"),col=c(4,3,5),lwd=2,inset=0.05)
```

# Comparing with survival values used in stage structured model

The MMIQT report considered stage stuctured models. Here we considered age stuctured models. Here we compare the two types of survival functions for a couple of extreme species. Since This comparison only makes sense by species/stock we look in particular at the species with the largest and shortest gestation duration.

For comparison, we overlay in the plots the stage-specific survival that was used in stage specific models that were referred to in DWH MMIQT 2015, Models and analyses for the quantification of injury to Gulf of Mexico cetaceans from the Deepwater Horizon Oil Spill, MM_TR.01 _Schwacke_Quantification.of.lnjury.to.GOM.Cetaceans.


## Sperm whales

```{r}
# Survival at age x
ages <- 0:200
plot(ages,px(ages*(SDF$gd[1]/SDF$gd[SDF$Species=="Pmac"]),pf[1,1],pf[1,2],pf[1,3],pf[1,4],pf[1,5],scaling=(SDF$gd[1]/SDF$gd[2])),type="l",lty=1,main="Survival probability at age x",ylab="Proportion",xlab="Age",col=4,ylim=c(0,1))
lines(ages,px(ages*(SDF$gd[1]/SDF$gd[SDF$Species=="Pmac"]),pm[1,1],pm[1,2],pm[1,3],pm[1,4],pm[1,5],scaling=(SDF$gd[1]/SDF$gd[2])),lty=2,col=4)
legend("bottomleft",legend=c("Females Pmac","Males Pmac","Stage structured values"),lty=c(1,2,2),col=c(4,4,1), inset=0.02,lwd=c(1,1,3))
#from Lance's files
PmacScalf <- 0.907
PmacSjuv <- 0.9424
PmacSFemale <- 0.9777
PmacSMale <- 0.918
#add lines
segments(x0=0,x1=2,y0=PmacScalf,y1=PmacScalf,lty=2,lwd=3)
segments(x0=2,x1=9,y0=PmacSjuv,y1=PmacSjuv,lty=2,lwd=3)
segments(x0=9,x1=100,y0=PmacSFemale,y1=PmacSFemale,lty=1,lwd=3)
segments(x0=9,x1=100,y0=PmacSMale,y1=PmacSMale,lty=2,lwd=3)
```

## Spinner dolphins


```{r}
# Survival at age x
ages <- 0:120
plot(ages,px(ages*(SDF$gd[1]/SDF$gd[SDF$Species=="Slon"]),pf[1,1],pf[1,2],pf[1,3],pf[1,4],pf[1,5],scaling=(SDF$gd[1]/SDF$gd[2])),type="l",lty=1,main="Survival probability at age x",ylab="Proportion",xlab="Age",col=5,ylim=c(0,1))
lines(ages,px(ages*(SDF$gd[1]/SDF$gd[SDF$Species=="Slon"]),pm[1,1],pm[1,2],pm[1,3],pm[1,4],pm[1,5],scaling=(SDF$gd[1]/SDF$gd[2])),lty=2,col=5)
legend("bottomleft",legend=c("Females Slon","Males Slon","Stage structured values"),lty=c(1,2,2),col=c(4,4,1), inset=0.02,lwd=c(1,1,3))
#from Lance's files
SlonScalf <- 0.85
SlonSjuv <- 0.95
SlonSFemale <- 0.97
SlonSMale <- 0.97
#add lines
segments(x0=0,x1=2,y0=SlonScalf,y1=SlonScalf,lty=2,lwd=3)
segments(x0=2,x1=9,y0=SlonSjuv,y1=SlonSjuv,lty=2,lwd=3)
segments(x0=9,x1=100,y0=SlonSFemale,y1=SlonSFemale,lty=1,lwd=3)
segments(x0=9,x1=100,y0=SlonSMale,y1=SlonSMale,lty=2,lwd=3)
```
