---
title: "Gestation duration"
author: ""
date: \today
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
  pdf_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

# Preamble

This is an Electronic Supplement to the manuscript Marques et al. "Quantifying Deepwater Horizon oil spill induced injury on pelagic cetaceans".

The master file containing links to all supplementary files related to this paper is this [Electronic Supplements master file](ElectronicSupplements.html).

If you make use of any of this material in your work, it would be appreciated if you would [contact Tiago
Marques](mailto:tiago.marques@st-andrews.ac.uk) to let him know.

# Background

Species-specific gestation duration was used as a scaling factor for reproduction and survival parameters. To obtain suitable scaling factor estimates, a review of grey and published literature was carried out to determine the most appropriate estimate of the length of gestation. Given the focus on the Gulf of Mexico, information close to the study region (e.g. East coast USA, Caribbean and Atlantic Ocean), where available, were favoured over estimates from other geographic regions (e.g. NW Pacific). Where no information was available to estimate gestation period for a species of interest, the value from closely related proxy species was used.

# Results

The best estimates of gestation duration for each taxonomic unit considered in the paper are shown below  (Table S4.1). Each section below describes the supporting information behind its selection or estimation. 

Note these are the values present in the column for gestation duration (gd) in the master file (SpeciesDefinitionFile4offshore.xlsx) that defines all parameters used in the simulations.

```{r}
library(knitr)
library(readxl)
temp <- read_excel("InputFiles/SpeciesDefinitionFile.xlsx",na = "NA")
Sps <- c("Bwsp", "Fatt", "Gmac", "Ggri", "Kosp", "Pele", "Pmac", "Satt", "Sbre", "Scly", "Scoe", "Sfro", "Slon", "Ttro", "Ttrs", "Ttru")
temp <- temp[temp$Species %in% Sps,]
temp <- temp[order(temp$Species),]
temp <- temp[,c("Species","latin","gd")]
kable(as.data.frame(temp),caption = "Gestation duration (gd), in days, for each of the taxonomic units considered",digits = 0)
```

# Justification for values considered

## Bottlenose dolphins (offshore and shelf stocks) 

Schwake, et al (in review) undertook a comprehensive review of the available data and literature on gestation duration in wild bottlenose dolphins and indicated a best estimate of 375 days We used that estimate in this study.  

## Beaked whales spp

New et al. (2013) reviewed the biology of a wide range of beaked whale species and provided a best estimate for most species of 12 months (citing Christensen (1973)). This is very likely a generalisation, but remains a best estimate given the paucity of data on beaked whale reproductive biology. While the much larger Baird’s beaked whale is estimated to have a gestation of approximately 17 months (Kasuya, 1977) from whaling records, we considered it is unlikely this applies to the beaked whale species sighted in the Gulf of Mexico (which are more in line with the best estimates of New et al (2013). We therefore used a gestation duration of 365 for Beaked whale spp. 

## Kogia spp.

A limited number of studies have estimated gestation in the Kogia spp. Much of the review literature (e.g. textbooks, summary papers) speculate that gestation length is possibly around 9.5 months (e.g. Ross, 1979;Pinedo, 1987;Willis and Baird, 1998) but few present the data to support this estimate.  The best empirical estimates for K. sima and K. breviceps available were in work by Plön (2004) suggesting gestation durations of 338 and 359 days (n: 25 and 26 females) respectively . Consequently we took the mean of the two best estimates, resulting in a gestation duration of 348 used in the models.

## Melon-headed whale 

The only published estimate of gestation period duration for this species is from Amano et al. (2014), based on a sample of 291 individuals from stranding events. This estimate is of 383 days and is broadly in line with more coarse estimates (Ridgway and Harrison, 1994;Bannister et al., 1996).

## Pygmy killer whale 

There are no data to inform this parameter in Pygmy killer whale. They are considered to be closest biologically to melon-headed whale, pilot whales and false killer whales (Baird, 2018), therefore we took the average  of these three gestation durations (383, 453 and 437 days (derived from Ferreira et al., 2014) respectively) as 424 days. 

## Risso’s dolphin 

The gestation duration has been estimated to be 13-14 months in the NW Pacific (Kasuya, 1985b)(estimates derived from 23 examined females) and 13.9 months in the Mediterranean Sea (based on 51 stranded specimens)(Raduán et al., 2007). We have used an estimate of 417 days averaging across these studies.

## Rough-toothed dolphin 

The reproductive biology of the rough-toothed dolphin species is poorly understood. Two females were in long-term human care in North America since 1996 and two females became sexually mature, resulting in four separate pregnancies. Mean gestation duration across the four pregnancies was 376 days (Staggs and Holmes, 2015).

## Short-finned pilot whale 

Limited information are available on the reproductive biology of short-finned pilot whales. Kasuya and Marsh (1984) provide an gestation duration estimate based on the examination of 373 females of 453 days (estimated as 14.9 months). In the absence of other estimates, we used that value in models. 

## Sperm whale 

A wide range of estimates are available for sperm whales globally. An exploration of geographic variations indicated a dimorphism between northern and southern hemisphere animals (slightly longer gestation in the southern hemisphere (Huggett and Widdas, 1951;Laws, 1959).  We took the average gestation across appropriate northern hemisphere studies where the rationale for calculations was available for assessment (Clarke, 1956;Laws, 1959;Chuzhakina, 1961;Ohsumi, 1965;Gambell, 1966). The resulting mean gestation duration estimate was 483 days. 

## Stenella dolphin species

Almost all information informing the reproductive biology of other *Stenella* spp. comes from the Eastern Tropical Pacific bycatch records and from animals taken in drive fisheries in the NW Pacific (e.g. Japan). The best estimates for pantropical spotted dolphin (*Stenella attenuata*) were of 342 (Kasuya, 1977), 345 (Perrin et al., 1976) and 349 days (Perrin and Hohn, 1994), resulting in an average of 345 days used for the species in this study.  Gestation duration in striped dolphins was estimated to be 365 days by Kasuya (1985a). For spinner dolphins (*Stenella longirostris*) estimates of 10.6 and 10.9 months (322 and 331 days) were available – and so the average was taken as 327 days (Perrin et al., 1977;Larese and Chivers, 2009). For Atlantic spotted dolphin, few records existed on their reproductive biology. Herzing (1997) suggested gestation would be similar to pantropical spotted dolphins and we use that value here (345 days). Similarly, no suitable records were found for the Clymene dolphin (*Stenella clymene*) so following Amaral et al. (2014) we took the average of striped and spinner dolphin gestation duration. This resulted in an estimated gestation duration of 346 days.

## Literature Cited

Amano, M., Yamada, T.K., Kuramochi, T., Hayano, A., Kazumi, A., and Sakai, T. (2014). Life history and group composition of melon‐headed whales based on mass strandings in Japan. Marine Mammal Science 30, 480-493.

Amaral, A.R., Lovewell, G., Coelho, M.M., Amato, G., and Rosenbaum, H.C. (2014). Hybrid speciation in a marine mammal: the clymene dolphin (*Stenella clymene*). PloS one 9, e83645.

Baird, R.W. (2018). "Pygmy Killer Whale: *Feresa attenuata*," in Encyclopedia of Marine Mammals. Elsevier), 788-790.

Bannister, J., Kemper, C.M., and Warneke, R.M. (1996). The action plan for Australian cetaceans. Australian Nature Conservation Agency Canberra, Australia.

Christensen, I. (1973). Age determination, age distribution and growth of bottlenose whales, *Hyperoodon ampullatus* (Forster), in the Labrador Sea. Norwegian journal of zoology 21, 331-340.

Chuzhakina, E. (1961). Morphological characterisation of the ovaries of the female sperm whale (*Physeter catodon* L. 1758) in connection with age determination. Trudy Inst. Morf. Zhivot 34, 33-53.

Clarke, R. (1956). Sperm Whales of the Azores, &c. University Press.
Ferreira, I.M., Kasuya, T., Marsh, H., and Best, P.B. (2014). False killer whales (*Pseudorca crassidens*) from Japan and South Africa: Differences in growth and reproduction. Marine Mammal Science 30, 64-84.

Gambell, R. (1966). Foetal growth and the breeding season of sperm whales. Norsk Hvalfangst-tidende 55, 113-118.

Herzing, D.L. (1997). The life history of free‐ranging Atlantic spotted dolphins (*Stenella frontalis*): age classes, color phases, and female reproduction. Marine Mammal Science 13, 576-595.

Huggett, A.S.G., and Widdas, W. (1951). The relationship between mammalian foetal weight and conception age. The Journal of Physiology 114, 306.

Kasuya, T. (1977). Age determination and growth of the Baird’s beaked whale with a comment on the fetal growth rate. Sci Rep Whales Res Inst 29, 1-20.

Kasuya, T. (1985a). Effect of exploitation on reproductive parameters of the spotted and striped dolphins off the Pacific coast of Japan. Scientific Reports of the Whales Research Institute 36, 107-138.

Kasuya, T. (1985b). Fishery-dolphin conflict in the Iki Island area of Japan. Marine mammals and fisheries, 253-272.

Kasuya, T., and Marsh, H. (1984). Life history and reproductive biology of the short-finned pilot whale, *Globicephala macrorhynchus*, off the Pacific coast of Japan. Report of the International Whaling Commission, Special 6, 259-310.

Larese, J.P., and Chivers, S.J. (2009). Growth and reproduction of female eastern and whitebelly spinner dolphins incidentally killed in the eastern tropical Pacific tuna purse-seine fishery. Canadian Journal of Zoology 87, 537-552.

Laws, R.M. (1959). The foetal growth rates of whales with special reference to the fin whale, *Balaenoptera physalus* Linn. Cambridge University Press.

New, L.F., Moretti, D.J., Hooker, S.K., Costa, D.P., and Simmons, S.E. (2013). Using energetic models to investigate the survival and reproduction of beaked whales (family Ziphiidae). PLoS ONE 8.

Ohsumi, S. (1965). Reproduction of the sperm whale in the North-West Pacific. Scientific Reports of the Whales Research Institute, Tokyo 19, 1-35.

Perrin, W., Holts, D., and Miller, R. (1977). Growth and reproduction of the eastern spinner dolphin, a geographical form of *Stenella longirostris* in the eastern tropical Pacific. Fishery bulletin 75, 725-750.

Perrin, W.F., Coe, J.M., and Zweifel, J.R. (1976). Growth and reproduction of the spotted porpoise, *Stenella attenuata*, in the offshore eastern tropical Pacific. Fishery Bulletin 74, 229-269.

Perrin, W.F., and Hohn, A.A. (1994). Pantropical spotted dolphin *Stenella attenuata*. Handbook of marine mammals 5, 71-98.

Pinedo, M.C. (1987). First record of a dwarf sperm whale from southwest Atlantic, with reference to osteology, food habits and reproduction. Sci. Rep. Whales Res. Inst 38, 171-186.

Plön, S. (2004). The status and natural history of pygmy (*Kogia breviceps*) and dwarf (*K. sima*) sperm whales off Southern Africa. Rhodes University Grahamstown, South Africa.

Raduán, A., Blanco, C., Fernández, M., and Raga, J. (Year). "Some aspects of the life history of the Risso’s dolphins *Grampus griseus* (Cuvier, 1812) in the western Mediterranean Sea", in: Proceedings of the Annual Conference of the European Cetacean Society).

Ridgway, S.H., and Harrison, R.J. (1994). The first book of dolphins. Academic Press.

Ross, G. (1979). Records of pygmy and dwarf sperm whales, genus *Kogia*, from southern Africa, with biological notes and some comparisons. Cape provincial Museums.

Staggs, L.A., and Holmes, S. (2015). "Reproductive Findings in Rough-Toothed Dolphins (*Steno bredanensis*) Maintained in Long-Term Human Care ", in: 46th Annual International Association for Aquatic Animal Medicine Meeting and Conference. (Chicago, IL. USA.: International Association for Aquatic Animal Medicine).

Willis, P.M., and Baird, R.W. (1998). Status of the dwarf sperm whale, *Kogia simus*, with special reference to Canada. Canadian Field-Naturalist 112, 114-125.







