---
title: "Scaling survival: comparing scaling based on gestation duration across species"
author: ""
date: \today
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
  pdf_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preamble

This is an Electronic Supplement to the manuscript Marques et al. "Quantifying Deepwater Horizon oil spill induced injury on pelagic cetaceans" submitted to Marine Ecology Progress Series (MEPS).

The master file containing links to all supplementary files related to this paper is [ES0_ElectronicSupplements](ES0_ElectronicSupplements.html).

If you make use of any of this material in your work, it would be appreciated if you would [contact Tiago
Marques](mailto:tiago.marques@st-andrews.ac.uk) to let him know.

## Version history

* 1.0 [date] Version included as an html Electronic supplement in the MEPS submission - *note to co-authors: this note will be deleted when we submit and we are not tracking versions prior to submitting to MEPS, that will be version 1.0 by definition* 

# Introduction

In this document we look at the age specific survival used for each species once scaled by the gestation duration, as described in the manuscript.

To understand why we ended up here a relevant source is Marques (2020) On scaling survival and the Siler model. A CARMMHA working document. Despite being a working document not prepared for a wider distribution, please contact tiago.marques@st-andrews.ac.uk if you are interested in a copy of said document.

# Reading relevant info

```{r,echo=FALSE}
library(readxl)
library(knitr)
library(ggplot2)
# Siler model functions are here
source("Functions/SilerFuns.R")       
```

We read in the species details

```{r}
#read the species definition file
SDF <- read_excel("InputFiles/SpeciesDefinitionFile.xlsx",na="NA")
```

and we set the species which we want results for, and then filter the input data for those species only.

```{r}
#remove irrelevant species rows
#SDF <- SDF[-c(1,6,7,14,16,22:27),]
sps<-c("Bwsp","Fatt","Gmac","Ggri","Kosp","Pele","Pmac","Satt","Sbre","Scly","Scoe","Sfro","Slon","Ttro","Ttrs","Ttru") 
SDF <- SDF[SDF$Species %in% sps,]
```

We read in *Tursiops truncatus* bottlenose dolphin (BND) Siler model posterior values. These can then be used to estimate:

1. survival conditional on age, and 
2. survivorship functions for this species. 

The corresponding values for all other species/stocks are obtained by scaling the *Tursiops truncatus* values by gestation duration.

```{r}
#--------------------------------------------------------------------------
# Survival
#--------------------------------------------------------------------------
# reading in Siler parameters, for males and females
# These files were provided by LS via email on the Fri 5/22/2020 4:10 PM
pf <- read.csv("InputFiles/var_female 15May2020.csv")
pm <- read.csv("InputFiles/var_male 15May2020.csv")
# remove useless ID first column 
pf<-pf[, -1]
pm<-pm[, -1]
# rename columns
names(pf) <- names(pm) <- c("a1", "a2", "a3", "b1", "b3", "rmean")
#--------------------------------------------------------------------------
```

# *Tursiops truncatus*

Here we represent the relevant functions and the associated variability is also illustrated.

## Survivorship function

Plotting the age specific survival aka survivorship function for BND. See Schwacke et al. 2017 for details on how this was derived based on a Siler model fitted to strandings data. 

```{r,cache=TRUE,fig.width = 10}
ages <- 0:70
ncols <- nrow(pf)
nrows <- length(ages)
#get a dataframe to use ggplot2
#objects to hold female and male realizations
lxFs <- matrix(NA,nrow=nrows,ncol=ncols)
lxMs <- matrix(NA,nrow=nrows,ncol=ncols)
#for each obseration of the posterior
for(i in 1:nrow(pf)){
  #get the funtion
  lxFs[,i] <- lx(ages,pf[i,1],pf[i,2],pf[i,3],pf[i,4],pf[i,5])
  lxMs[,i] <- lx(ages,pm[i,1],pm[i,2],pm[i,3],pm[i,4],pm[i,5])
}
#arrange as data frame
lxFs2 <- data.frame(ages=ages,lxFs)
lxMs2 <- data.frame(ages=ages,lxMs)
# add means and relevant quantiles
lxFs2$mean=rowMeans(lxFs)
lxMs2$mean=rowMeans(lxMs)
lxFs2$q025=apply(X=lxFs,MARGIN = 1,FUN=quantile,probs=0.025)
lxMs2$q025=apply(X=lxMs,MARGIN = 1,FUN=quantile,probs=0.025)
lxFs2$q975=apply(X=lxFs,MARGIN = 1,FUN=quantile,probs=0.975)
lxMs2$q975=apply(X=lxMs,MARGIN = 1,FUN=quantile,probs=0.975)
#add the sex
lxFs2$sex="Female"
lxMs2$sex="Male"
#make single object
lxMF <- rbind(lxFs2,lxMs2)
#plot
#https://www.r-graph-gallery.com/104-plot-lines-with-error-envelopes-ggplot2.html
ggplot(data=lxMF, aes(x=ages, y=mean, ymin=q025, ymax=q975, fill=sex, linetype=sex)) + geom_line() + 
 geom_ribbon(alpha=0.5) + 
 xlab("Age") + 
 ylab("Proportion surviving")
```

## Age specific Survival

```{r,cache=TRUE,fig.width = 10}
ages <- 0:70
ncols <- nrow(pf)
nrows <- length(ages)
#get a dataframe to use ggplot2
#objects to hold female and male realizations
pxFs <- matrix(NA,nrow=nrows,ncol=ncols)
pxMs <- matrix(NA,nrow=nrows,ncol=ncols)
#for each obseration of the posterior
for(i in 1:nrow(pf)){
  #get the funtion
  pxFs[,i] <- px(ages,pf[i,1],pf[i,2],pf[i,3],pf[i,4],pf[i,5])
  pxMs[,i] <- px(ages,pm[i,1],pm[i,2],pm[i,3],pm[i,4],pm[i,5])
}
#arrange as data frame
pxFs2 <- data.frame(ages=ages,pxFs)
pxMs2 <- data.frame(ages=ages,pxMs)
# add means and relevant quantiles
pxFs2$mean=rowMeans(pxFs)
pxMs2$mean=rowMeans(pxMs)
pxFs2$q025=apply(X=pxFs,MARGIN = 1,FUN=quantile,probs=0.025)
#two out of 4000*71 values are NA's?
pxMs2$q025=apply(X=pxMs,MARGIN = 1,FUN=quantile,probs=0.025,na.rm=TRUE)
pxFs2$q975=apply(X=pxFs,MARGIN = 1,FUN=quantile,probs=0.975)
pxMs2$q975=apply(X=pxMs,MARGIN = 1,FUN=quantile,probs=0.975,na.rm=TRUE)
#add the sex
pxFs2$sex="Female"
pxMs2$sex="Male"
#make single object
pxMF <- rbind(pxFs2,pxMs2)
#plot
#https://www.r-graph-gallery.com/104-plot-lines-with-error-envelopes-ggplot2.html
ggplot(data=pxMF, aes(x=ages, y=mean, ymin=q025, ymax=q975, fill=sex, linetype=sex)) + geom_line() + 
 geom_ribbon(alpha=0.5) + 
 xlab("Age") + 
 ylab("Age specific Survival")
```
It is relevant to appreciate that the survival curves across iterations will tend to cross, and therefore we cannot really talk about higher survival probabilities across the board for a single iteration, the survival might be higher in iteration i for younger animals but higher for older animals. This is illustrated in the image below, where a couple of iterations are highlighted (in green and blue) against all the iterations.

```{r, fig.width = 10,fig.height=6}
par(mfrow=c(1,2))
set.seed(123)
#get the funtion
plot(c(0,72),c(0,1),type="n",main="Female survival",xlab="Age (in years)",ylab="Survival")
for(i in 2:(nrow(pf)-4)){
    lines(1:71,pxFs2[,i])
}

lines(1:71,pxFs2[,23],lwd=3,col="green")

lines(1:71,pxFs2[,32],lwd=3,col="blue")

plot(c(0,72),c(0,1),type="n",main="Male survival",xlab="Age (in years)",ylab="Survival")
for(i in 2:(nrow(pf)-4)){
  lines(1:71,pxMs2[,i])
}

lines(1:71,pxMs2[,23],lwd=3,col="green")

lines(1:71,pxMs2[,32],lwd=3,col="blue")
```


# Comparing all species

The scaling is done by gestation duration, so we look at the gestation duration considered for each of the species. The rationale for the gestation duration used is provided on a separate document included as a separate Electronic Supplement (ElectronicSupplement4_GD).

The range in gestation duration is shown below (note the x axis is just an index, species are shown from left to right in the increasing order of their gestation duration).

```{r, fig.width = 10, fig.height = 6}
par(mfrow = c(1, 1),mar=c(0.2,4,0.2,0.2))
plot(1:16, sort(SDF$gd), ylab = "Gestation duration (in days)", xlab = "", type = "n",xaxt="n",ylim=c(0,max(SDF$gd)+10))
text(1:16, sort(SDF$gd), labels = SDF$Species[order(SDF$gd)])
```

## Survivorship function

We present the scalled survivorship by function below. For reference BB BND shown in green, sperm whale in blue and spinner dolphin in light blue

```{r, fig.width = 10,fig.height=6}
par(mfrow = c(2, 1),mar = c(4, 4, 0.5, 0.5))
ages <- 0:70
# Proportion surviving to age x
plot(ages, lx(ages*1, pf[1, 1], pf[1, 2], pf[1, 3], pf[1, 4], pf[1, 5]), type = "l", main = "", ylab = "Proportion", xlab = "Age (females)", lty = 1)
for(i in 1:nrow(SDF)){
  # Proportion surviving to age x
lines(ages, lx(ages*(SDF$gd[1]/SDF$gd[i]), pf[1, 1], pf[1, 2], pf[1, 3], pf[1, 4], pf[1, 5]), type="l", main="Proportion surviving to age x", ylab="Proportion", xlab="Age (females)", lty=i,col=ifelse(SDF$Species[i]=="Ttru",3,ifelse(SDF$Species[i]=="Pmac",4,ifelse(SDF$Species[i]=="Slon",5,1))),lwd=ifelse(SDF$Species[i]=="Ttru",2,ifelse(SDF$Species[i]=="Pmac",2,ifelse(SDF$Species[i]=="Slon",2,1))))
}
plot(ages, lx(ages*1 , pm[1, 1], pm[1, 2], pm[1, 3], pm[1, 4], pm[1, 5]), type = "l", lty = 1, xlab="Age (males)", ylab="Proportion")
for(i in 1:nrow(SDF)){
# Proportion surviving to age x
lines(ages, lx(ages*(SDF$gd[1]/SDF$gd[i]), pm[1, 1], pm[1, 2], pm[1, 3], pm[1, 4], pm[1, 5]), lty=i, xlab="Age (males)",col=ifelse(SDF$Species[i]=="Ttru",3,ifelse(SDF$Species[i]=="Pmac",4,ifelse(SDF$Species[i]=="Slon",5,1))),lwd=ifelse(SDF$Species[i]=="Ttru",2,ifelse(SDF$Species[i]=="Pmac",2,ifelse(SDF$Species[i]=="Slon",2,1))))
}
legend("topright",legend=c("Pmac","Ttru","Slon"),col=c(4,3,5),lwd=2,inset=0.05)
```

## Age specific survival

We present the scalled age specific survival by function below. For reference BB BND shown in green, sperm whale in blue and spinner dolphin in light blue

```{r, fig.width = 10,fig.height=6}
par(mfrow = c(2, 1),mar = c(4, 4, 0.5, 0.5))
ages <- 0:140
# Proportion surviving to age x
plot(ages, px(ages*1, pf[1, 1], pf[1, 2], pf[1, 3], pf[1, 4], pf[1, 5],scaling=1), type = "l", main = "", ylab = "Survival", xlab = "Age (females)", lty = 1, ylim = c(0, 1))
for(i in 1:nrow(SDF)){
  # Proportion surviving to age x
lines(ages, px(ages*(SDF$gd[1]/SDF$gd[i]), pf[1, 1], pf[1, 2], pf[1, 3], pf[1, 4], pf[1, 5],scaling=(SDF$gd[1]/SDF$gd[i])), type="l", main="Proportion surviving to age x", ylab="Proportion", xlab="Age (females)", lty=i,col=ifelse(SDF$Species[i]=="Ttru",3,ifelse(SDF$Species[i]=="Pmac",4,ifelse(SDF$Species[i]=="Slon",5,1))),lwd=ifelse(SDF$Species[i]=="Ttru",2,ifelse(SDF$Species[i]=="Pmac",2,ifelse(SDF$Species[i]=="Slon",2,1))))
}
plot(ages, px(ages*1 , pm[1, 1], pm[1, 2], pm[1, 3], pm[1, 4], pm[1, 5],scaling=1), type = "l", lty = 1, xlab="Age (males)", ylab="Survival", ylim = c(0, 1))
for(i in 1:nrow(SDF)){
# Proportion surviving to age x
lines(ages, px(ages*(SDF$gd[1]/SDF$gd[i]), pm[1, 1], pm[1, 2], pm[1, 3], pm[1, 4], pm[1, 5],scaling=(SDF$gd[1]/SDF$gd[i])), lty=i, xlab="Age (males)",col=ifelse(SDF$Species[i]=="Ttru",3,ifelse(SDF$Species[i]=="Pmac",4,ifelse(SDF$Species[i]=="Slon",5,1))),lwd=ifelse(SDF$Species[i]=="Ttru",2,ifelse(SDF$Species[i]=="Pmac",2,ifelse(SDF$Species[i]=="Slon",2,1))))
}
legend("topright",legend=c("Pmac","Ttru","Slon"),col=c(4,3,5),lwd=2,inset=0.05)
```

# Comparing with survival values used in stage structured model

The [MMIQT report](https://www.fws.gov/doiddata/dwh-ar-documents/876/DWH-AR0105866.pdf) considered stage structured models. In this paper we considered age structured models. Here we compare the two types of survival functions for a couple of extreme species. Since This comparison only makes sense by species/stock we look in particular at the species with the largest and shortest gestation duration, namely the sperm whale and the spinner dolphin.For comparison, we overlay in the plots our survival by age and the stage-specific survival that was used in stage specific models that were referred to in DWH MMIQT 2015.


## Sperm whales

```{r}
# Survival at age x
ages <- 0:200
plot(ages,px(ages*(SDF$gd[1]/SDF$gd[SDF$Species=="Pmac"]),pf[1,1],pf[1,2],pf[1,3],pf[1,4],pf[1,5],scaling=(SDF$gd[1]/SDF$gd[2])),type="l",lty=1,main="Survival probability at age x",ylab="Proportion",xlab="Age",col=4,ylim=c(0,1))
lines(ages,px(ages*(SDF$gd[1]/SDF$gd[SDF$Species=="Pmac"]),pm[1,1],pm[1,2],pm[1,3],pm[1,4],pm[1,5],scaling=(SDF$gd[1]/SDF$gd[2])),lty=2,col=4)
legend("bottomleft",legend=c("Females Pmac","Males Pmac","Stage structured values"),lty=c(1,2,2),col=c(4,4,1), inset=0.02,lwd=c(1,1,3))
#from Lance's files
PmacScalf <- 0.907
PmacSjuv <- 0.9424
PmacSFemale <- 0.9777
PmacSMale <- 0.918
#add lines
segments(x0=0,x1=2,y0=PmacScalf,y1=PmacScalf,lty=2,lwd=3)
segments(x0=2,x1=9,y0=PmacSjuv,y1=PmacSjuv,lty=2,lwd=3)
segments(x0=9,x1=100,y0=PmacSFemale,y1=PmacSFemale,lty=1,lwd=3)
segments(x0=9,x1=100,y0=PmacSMale,y1=PmacSMale,lty=2,lwd=3)
```

## Spinner dolphins


```{r}
# Survival at age x
ages <- 0:120
plot(ages,px(ages*(SDF$gd[1]/SDF$gd[SDF$Species=="Slon"]),pf[1,1],pf[1,2],pf[1,3],pf[1,4],pf[1,5],scaling=(SDF$gd[1]/SDF$gd[2])),type="l",lty=1,main="Survival probability at age x",ylab="Proportion",xlab="Age",col=5,ylim=c(0,1))
lines(ages,px(ages*(SDF$gd[1]/SDF$gd[SDF$Species=="Slon"]),pm[1,1],pm[1,2],pm[1,3],pm[1,4],pm[1,5],scaling=(SDF$gd[1]/SDF$gd[2])),lty=2,col=5)
legend("bottomleft",legend=c("Females Slon","Males Slon","Stage structured values"),lty=c(1,2,2),col=c(4,4,1), inset=0.02,lwd=c(1,1,3))
#from Lance's files
SlonScalf <- 0.85
SlonSjuv <- 0.95
SlonSFemale <- 0.97
SlonSMale <- 0.97
#add lines
segments(x0=0,x1=2,y0=SlonScalf,y1=SlonScalf,lty=2,lwd=3)
segments(x0=2,x1=9,y0=SlonSjuv,y1=SlonSjuv,lty=2,lwd=3)
segments(x0=9,x1=100,y0=SlonSFemale,y1=SlonSFemale,lty=1,lwd=3)
segments(x0=9,x1=100,y0=SlonSMale,y1=SlonSMale,lty=2,lwd=3)
```


# References

DWH MMIQT 2015, Models and analyses for the quantification of injury to Gulf of Mexico
cetaceans from the Deepwater Horizon Oil Spill, MM_TR.01_Schwacke_Quantification.of.lnjury.to.GOM.Cetaceans [LINK](https://www.fws.gov/doiddata/dwh-ar-documents/876/DWH-AR0105866.pdf).

Schwacke, L.H., L. Thomas, R.S. Wells, W.E. McFee, A.A. Hahn, K.D. Mullin, E.S. Zolman, B.M. Quigley, T.K. Rowles and J.H. Schwacke. 2017. An age-, sex- and class-structured population model for estimating nearshore common bottlenose dolphin injury following the Deepwater Horizon oil spill. *Endangered Species Research* **33**: 265-279. DOI: [10.3354/esr00777](https://doi.org/10.3354/esr00777).