---
title: "Assessing injury on GOM cetaceans"
author: ""
date: \today
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
  pdf_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
#to NOT disply scientific notation
options(scipen=999)
```


# Preamble

This is an Electronic Supplement to the manuscript Marques et al. "Quantifying Deepwater Horizon oil spill induced injury on pelagic cetaceans" submitted to Marine Ecology Progress Series (MEPS).

The master file containing links to all supplementary files related to this paper is [ES0_ElectronicSupplements](ES0_ElectronicSupplements.html).

If you make use of any of this material in your work, it would be appreciated if you would [contact Tiago
Marques](mailto:tiago.marques@st-andrews.ac.uk) to let him know.

## Version history

* 1.0 [date] Version included as an html Electronic supplement in the MEPS submission - *note to co-authors: this note will be deleted when we submit and we are not tracking versions prior to submitting to MEPS, that will be version 1.0 by definition* 

# Introduction

In this document we present the results of the age, sex and class structured model implemented for all species/stock in the GOM. 

We report here the injury metrics originally considered in Schwacke et al. (2017): (1) lost cetacean years (LCY), the difference between the baseline and injured population sizes, summed over the entire modeled time period; (2) years to recovery (YTR), the number of years required before the injured population trajectory reaches 95% of the baseline population trajectory; and (3) maximum proportional decrease (MPD), the difference between the 2 population trajectories when the injured trajectory is at its lowest point, divided by the baseline.  Note that LCY is intuitively the metric that is most dependent on initial population size.

# Running the simulations

The code below runs the simulations for all species, and stores the results in appropriate files, stored in appropriate species specific folders. The code below is for completeness only, and IS NOT, AND SHOULD NOT BE, run when this dynamic report is produced.

```{r}
#define the species we run simulations for
Sps <- c("Bwsp", "Fatt", "Ggri", "Gmac", "Kosp", "Pele", "Pmac", "Satt", "Sbre", "Scly", "Scoe", "Sfro", "Slon", "Ttro", "Ttrs")
#define the number of iterations to run
ns <- 10000
# define the number of years each iteration is run for
ny <- 75
```

```{r getresults,eval=FALSE}
# this runs all the simulations and produces outputs for all species
# this code chuck is a separate file: "RunAllSims.R"
# that is the file that should be sourced to run the code
# this code is therefore here just for completeness
# DON NOT turn the "eval=FALSE" into "eval=TRUE"
# unless you are willing to wait for many (many!) hours 
# (or even days, depending on the "ns" you choose)
# for your dynamic report to compile
source("Functions/reqfuns.R")         # most functions are here
source("Functions/SilerFuns.R")       # Siler model functions are here
source("Functions/runSims.R")         # To run the simulations for all species
for (i in Sps){
  runPopSims(Sp = i, nsims = ns, nyears = ny, seed = 7134672)
}
```


```{r,cache=FALSE}
source("Functions/reqfuns.R")         # most functions are here
```

Creating objects to hold summary results:

```{r,cache=FALSE}
# Set up a table to hold summary statistics by species
sumres <- data.frame(Sp=Sps, No=NA, peo=NA, LCYo=NA, MPDo=NA, YTRo=NA, Nstart=NA, Nexp=NA, Pexp=NA, LCY=NA, MPD=NA, YTR=NA,  LCYmed=NA, MPDmed=NA, YTRmed=NA)
# Injury metrics and respective 95% quantile-method CI's 
injuryCI <- data.frame(Sp=Sps, LCYmed = NA, LCYl = NA, LCYu = NA, MPDmed = NA, MPDl = NA, MPDu = NA, YTRmed = NA, YTRl = NA, YTRu = NA)
# Some relevant inputs (gd gestation duration and sr survival reduction)
inputs <- data.frame(Sp=Sps, gd = NA, sr = NA)
```

The taxonomic units and the corresponding codes considered in this document are:

* Bwsp beaked whales	Beaked whales spp
* Fatt pygmy killer whale	*Feresa attenuata*
* Ggri Risso's dolphin	*Grampus griseus*
* Gmac short-finned pilot whale	*Globicephala macrorhynchus*
* Kosp kogia	*Kogia* sp.
* Pele melon-headed whale	*Peponocephala electra*
* Pmac sperm whale	*Physeter macrocephalus*
* Satt pantropical spotted dolphin *Stenella attenuata*
* Sbre rough-toothed dolphin	*Steno bredanensis*
* Scly Clymene dolphin	*Stenella clymene*
* Scoe striped dolphin	*Stenella coeruleoalba*
* Sfro Atlantic spotted dolphin	*Stenella frontalis*
* Slon spinner dolphin	*Stenella longirostris*
* Ttro offshore bottlenose dolphins	*Tursiops truncatus*
* Ttrs shelf bottlenose dolphins *Tursiops truncatus*

We explicitly re-order the results tables upfront so that results are organized by alphabetical order of the 4 letter code used to describe the taxonomic unit (this is also the order that the species are reported in tables 1 to 3 of the offshore paper).

```{r}
# note there is no effect if Sps is already in alphabetical order
# but it is here to make sure that this code will work even if
# the vector Sps is changed
injuryCI <- injuryCI[order(Sps),]
sumres <- sumres[order(Sps),]
inputs <- inputs[order(Sps),]
```

Note that the initial population sizes and proportion exposed shown below are obtained after removing the very rare instances where the population size would be considered too large to be reasonable. This is imported from a `txt` file produced inside [IniPopSizes](IniPopSizes.html). See [IniPopSizes](IniPopSizes.html) and the main paper for details on this choice.

```{r}
IniPopSizes <- read.table(file = "Table4GetInjury.txt",header=TRUE)
IniPopSizes <- IniPopSizes[order(Sps),]
```

# Results by species

We represent in turn, for each species, the simulated population trajectories under the oil spill and under the baseline scenarios, as well as the histograms of the distributions of the 3 injury metrics for each species.

## *Physeter macrocephalus* (Pmac, sperm whale)

```{r,cache=FALSE}
gc();Sp <- "Pmac"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd 
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r,cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```

and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## *Kogia* sp. (Kosp, kogia)

```{r,cache=FALSE}
gc();Sp <- "Kosp"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r, cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```

and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## *Stenella clymene* (Scly, Clymene dolphin)	

```{r,cache=FALSE}
gc();Sp <- "Scly"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r,cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```
and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## *Peponocephala electra* (Pele, melon-headed whale)	

```{r,cache=FALSE}
gc();Sp <- "Pele"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r,cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```

and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## *Tursiops truncatus* (Ttro, offshore bottlenose dolphin)	

```{r,cache=FALSE}
gc();Sp <- "Ttro"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r,cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```

and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## *Globicephala macrorhynchus* (Gmac, short finned pilot whale) 

```{r,cache=FALSE}
gc();Sp <- "Gmac"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r,cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```

and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## *Feresa attenuata* (Fatt, pygmy killer whale)	


```{r,cache=FALSE}
gc();Sp <- "Fatt"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r,cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```

and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## *Grampus griseus* (Ggri, Risso's dolphin)	

```{r}
gc();Sp <- "Ggri"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r}
plotSims(simres)
```

Injury measures

```{r}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```

and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## *Steno bredanensis* (Sbre, rough-toothed dolphin)	

```{r,cache=FALSE}
gc();Sp <- "Sbre"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r,cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```

and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## *Stenella attenuata* (Satt, pantropical spotted dolphin)	

```{r,cache=FALSE}
gc();Sp <- "Satt"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r,cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```

and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## *Tursiops truncatus* (Ttrs, shelf bottlenose dolphin)	

```{r,cache=FALSE}
gc();Sp <- "Ttrs"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r,cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```

and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## *Stenella longirostris* (Slon, spinner dolphin)	

```{r,cache=FALSE}
gc();Sp <- "Slon"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r,cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```

and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## *Stenella coeruleoalba* (Scoe, striped dolphin)	


```{r,cache=FALSE}
gc();Sp <- "Scoe"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r,cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```


and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## *Stenella frontalis* (Sfro, Atlantic spotted dolphin)	


```{r,cache=FALSE}
gc();Sp <- "Sfro"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r,cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```

and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

## Beaked whales spp (Bwsp)
 
```{r,cache=FALSE}
gc();Sp <- "Bwsp"
SpInfo <- getSpData(Sp)
load(file=paste0("InOutBySp/", SpInfo$folder, "/", Sp, "simres", ns, "Sim.RData"))
inputs$gd[inputs$Sp==Sp] <- SpInfo$gd
inputs$sr[inputs$Sp==Sp] <- with(SpInfo, mean(rbeta(10000, sra, srb)*(sru-srl)+srl))
```

Population trajectories

```{r,cache=FALSE}
plotSims(simres)
```

Injury measures

```{r,cache=FALSE}
x <- getInjury(simres)
colMeans(injury)
apply(injury, 2, median)
```

Collect information to summary table

```{r,cache=FALSE}
sumres[sumres$Sp==Sp, 2:15]<-c(SpInfo$N, SpInfo$pe, SpInfo$LCY, SpInfo$MPD, SpInfo$YTR, IniPopSizes$N[IniPopSizes$Sp==Sp], IniPopSizes$Ne[IniPopSizes$Sp==Sp], IniPopSizes$Pe[IniPopSizes$Sp==Sp], mean(injury$LCY), mean(injury$MPD), mean(injury$YTR),  median(injury$LCY), median(injury$MPD), median(injury$YTR))
```

and collect information to injury summary table

```{r}
injuryCI[injuryCI$Sp==Sp, 2:10]<-c(median(injury$LCY), quantile(injury$LCY, probs=c(0.025,0.975)), median(injury$MPD), quantile(injury$MPD, probs = c(0.025, 0.975)), median(injury$YTR), quantile(injury$YTR,probs = c(0.025, 0.975)))
```

# Comparisons across species

In this section we present some summary analysis o the results across the 15 taxonomic units considered.

```{r}
#just some rounding for nicer visualization below
sumres$Nstart=round(sumres$Nstart)
sumres$Pexp=round(sumres$Pexp,3)
```

## Initial population and proportion exposed

The population size and the proportion exposed are $a priori$ expected to be key determinants of injury, since the population size will have a direct impact on LCY and the proportion exposed should have a direct impact on all 3 measures.

```{r,cache=FALSE}
kable(sumres[, c(1, 2, 7, 3, 9)],caption = "Initial population size (N) and proportion exposed (pe or Pexp) for the different stocks considered.  The suffix o indicates old results from the MMIQT report and Schwacke et al. 2017",digits = c(0, 0, 0, 3, 3))
```

The proportion exposed against the initial population size is shown here for the 15 taxonomic units

```{r}
plot(sumres[, c(7,9)],xlab="Initial population size",ylab="Proportion exposed",type="n")
text(x = sumres[, 7],y=sumres[, 9],labels = sumres[, 1])
```


***DELETE IMAGES BELOW AND REMOVE MMIQT RESULTS ABOVE?***

The image below illustrates that there is no clear relationship between the initial population size assumed before (DWH MMIQT 2015) and the one being currently considered. This was expected given the completely different way in which population sizes were obtained in this exercise (via a spatial model) and before (a coarser stratification approach).

```{r,cache=FALSE}
with(sumres, plot(No,Nstart))
abline(0, 1)
```

The same happens for the proportion exposed, for the same reasons.

```{r,cache=FALSE}
with(sumres, plot(peo, Pexp))
abline(0,1)
```

This illustrates how direct comparisons of results between here and the previous DWH MMIQT (2015) results is far from straightforward, since not only the model has changed, but the inputs to the model have also changed.

***DELETE IMAGES ABOVE AND REMOVE MMIQT RESULTS ABOVE?***

## Injury metrics

### Means and Medians

In the following table we present the means and medians of the injury metrics for the current simulations, while the previous DWH MMIQT (2015) results are also shown for comparison:

```{r,cache=FALSE}
sumres$MPD <- -(sumres$MPD/100)
kable(sumres[, c(1, 4, 10, 13, 5, 11, 6, 12)],caption = "Injury results for the different stocks considered. The suffix o indicates old results MMIQT, med means median",digits = c(0, 0, 0, 0, 3, 3, 1, 1))
```

### Precision

In this following table we present also the 95% confidence intervals for our current estimates of injury. Note this corresponds to the results in table 3 in the main paper.

```{r}
kable(injuryCI, caption = "Median injury results for the different stocks considered and respective 95% confidence intervals by the percentile method",digits = c(0, 0, 0, 0, 3, 3, 3, 1, 1, 1))
```

Producing the values that are used in table 3 of the paper

```{r}
Table3LCY <-  paste0(round(injuryCI$LCYmed,0), " (",round(injuryCI$LCYl,0),", ",round(injuryCI$LCYu),") ")
Table3MPD <-  paste0(round(injuryCI$MPDmed,1), " (",round(injuryCI$MPDl,1),", ",round(injuryCI$MPDu,1), ") ")
Table3YTR <-  paste0(round(injuryCI$YTRmed,1), " (",round(injuryCI$YTRl,1)," ,",round(injuryCI$YTRu,1),")")
# IMPORTANT
# note removing row 16, Ttru, not for the pelagic paper
# These are outputed as 3 separate tables as 
# that facilitates copy-pasting into
# the main paper word document
write.table(Table3LCY[-16],"Table3LCY.txt",sep="\t",row.names = FALSE,quote = FALSE,col.names = FALSE)
write.table(Table3MPD[-16],"Table3MPD.txt",sep="\t",row.names = FALSE,quote = FALSE,col.names = FALSE)
write.table(Table3YTR[-16],"Table3YTR.txt",sep="\t",row.names = FALSE,quote = FALSE,col.names = FALSE)
```


```{r}
#creating output tables for others to use
write.table(injuryCI,"injuryCI.txt")
write.table(sumres,"sumres.txt")
```

We can also represent the above information visually

```{r}
par(mfrow=c(1,1),mar=c(4,4,0.5,0.5))
plot(sort(injuryCI$LCYmed),1:15,xlab="Lost Cetacean Years",ylab="Taxonomic unit",yaxt="n",xlim=c(0,72000))
order.index<-order(injuryCI$LCYmed)
axis(2,at=1:15,labels=injuryCI$Sp[order.index],las=2)
for(i in 1:15){
  segments(x0=injuryCI$LCYl[i],y0=(1:15)[order.index[i]],x1=injuryCI$LCYu[i],y1=(1:15)[order.index[i]])
}
abline(v=0,lty=2)
```


# Discussion (work in progress)

To facilitate the discussion of the results obtained we present below explicitly comparisons between the previous values of injury (DWH MMIQT , 2015) and the current values for injury we are estimating.

We present such comparisons first for LCY

```{r,cache=FALSE}
with(sumres,plot(LCYo,LCY,type="n"))
with(sumres,text(LCYo,LCY,Sp))
abline(0,1,lty=2)
#text(70000,2000,"Kogia")
#text(75000,15000,"Shelf dolphins")
#text(120000,3000,"rough-toothed dolphins")
#text(46000,36000,"BB BND")
```

followed by MPD

```{r}
with(sumres,plot(MPDo,MPD,type="n"))
with(sumres,text(MPDo,MPD,Sp))
abline(0, 1, lty = 2)
```

and finally YTR

```{r,cache=FALSE}
with(sumres,plot(YTRo,YTR,type="n"))
with(sumres,text(YTRo,YTR,Sp))
abline(0, 1, lty = 2)
```

We also present the values of the injury metrics as a function of the key parameters going in the model:

the proportion exposed

```{r}
#if you uncomment several lines  below you see the plot contrasting with DWH MMIQT (2015) results
#LCY
plot(sumres$Pexp,sumres$LCY, type="n",col=3, xlab="P(exposed)",ylab="LCY")
#plot(c(sumres$peo,sumres$Pexp), c(sumres$LCYo,sumres$LCY), type="n",col=3, xlab="P(exposed)",ylab="LCY")
#points(sumres$peo, sumres$LCYo, col=3)
#abline(lm(sumres$LCYo~sumres$peo),col=3,lty=2)
points(sumres$Pexp, sumres$LCY,col=4)
abline(lm(sumres$LCY~sumres$Pexp),col=4,lty=2)
#legend("topright",col=3:4,lty=2,c("old","new"),inset=0.05)
#MPD
plot(sumres$Pexp, sumres$MPD, type="n",col=3, xlab="P(exposed)",ylab="MPD")
#plot(c(sumres$peo,sumres$Pexp), c(sumres$MPDo,sumres$MPD), type="n",col=3, xlab="P(exposed)",ylab="MPD")
#points(sumres$peo, sumres$MPDo, col=3)
#abline(lm(sumres$MPDo~sumres$peo),col=3,lty=2)
points(sumres$Pexp, sumres$MPD,col=4)
abline(lm(sumres$MPD~sumres$Pexp),col=4,lty=2)
#legend("topright",col=3:4,lty=2,c("old","new"),inset=0.05)
#YTR
plot(sumres$Pexp, sumres$YTR, type="n",col=3, xlab="P(exposed)",ylab="YTR")
#plot(c(sumres$peo,sumres$Pexp), c(sumres$YTRo,sumres$YTR), type="n",col=3, xlab="P(exposed)",ylab="YTR")
#points(sumres$peo, sumres$YTRo, col=3)
#abline(lm(sumres$YTRo~sumres$peo),col=3,lty=2)
points(sumres$Pexp, sumres$YTR,col=4)
abline(lm(sumres$YTR~sumres$Pexp),col=4,lty=2)
#legend("topright",col=3:4,lty=2,c("old","new"),inset=0.05)
```

initial population size

```{r}
#if you uncomment several lines  below you see the plot contrasting with DWH MMIQT (2015) results
#LCY
plot(sumres$Nstart,sumres$LCY, type="n",col=3, xlab="initial population size",ylab="LCY")
#plot(c(sumres$peo,sumres$Pexp), c(sumres$LCYo,sumres$LCY), type="n",col=3, xlab="P(exposed)",ylab="LCY")
#points(sumres$peo, sumres$LCYo, col=3)
#abline(lm(sumres$LCYo~sumres$peo),col=3,lty=2)
points(sumres$Nstart, sumres$LCY,col=4)
abline(lm(sumres$LCY~sumres$Nstart),col=4,lty=2)
#legend("topright",col=3:4,lty=2,c("old","new"),inset=0.05)
#MPD
plot(sumres$Nstart, sumres$MPD, type="n",col=3, xlab="initial population size",ylab="MPD")
#plot(c(sumres$peo,sumres$Pexp), c(sumres$MPDo,sumres$MPD), type="n",col=3, xlab="P(exposed)",ylab="MPD")
#points(sumres$peo, sumres$MPDo, col=3)
#abline(lm(sumres$MPDo~sumres$peo),col=3,lty=2)
points(sumres$Nstart, sumres$MPD,col=4)
abline(lm(sumres$MPD~sumres$Nstart),col=4,lty=2)
#legend("topright",col=3:4,lty=2,c("old","new"),inset=0.05)
#YTR
plot(sumres$Nstart, sumres$YTR, type="n",col=3, xlab="initial population size",ylab="YTR")
#plot(c(sumres$peo,sumres$Pexp), c(sumres$YTRo,sumres$YTR), type="n",col=3, xlab="P(exposed)",ylab="YTR")
#points(sumres$peo, sumres$YTRo, col=3)
#abline(lm(sumres$YTRo~sumres$peo),col=3,lty=2)
points(sumres$Nstart, sumres$YTR,col=4)
abline(lm(sumres$YTR~sumres$Nstart),col=4,lty=2)
#legend("topright",col=3:4,lty=2,c("old","new"),inset=0.05)
```

gestation duration

```{r}
#if you uncomment several lines  below you see the plot contrasting with DWH MMIQT (2015) results
#LCY
plot(inputs$gd,sumres$LCY, type="n",col=3, xlab="gestation duration",ylab="LCY")
#plot(c(sumres$peo,sumres$Pexp), c(sumres$LCYo,sumres$LCY), type="n",col=3, xlab="P(exposed)",ylab="LCY")
#points(sumres$peo, sumres$LCYo, col=3)
#abline(lm(sumres$LCYo~sumres$peo),col=3,lty=2)
points(inputs$gd, sumres$LCY,col=4)
abline(lm(sumres$LCY~inputs$gd),col=4,lty=2)
#legend("topright",col=3:4,lty=2,c("old","new"),inset=0.05)
#MPD
plot(inputs$gd, sumres$MPD, type="n",col=3, xlab="gestation duration",ylab="MPD")
#plot(c(sumres$peo,sumres$Pexp), c(sumres$MPDo,sumres$MPD), type="n",col=3, xlab="P(exposed)",ylab="MPD")
#points(sumres$peo, sumres$MPDo, col=3)
#abline(lm(sumres$MPDo~sumres$peo),col=3,lty=2)
points(inputs$gd, sumres$MPD,col=4)
abline(lm(sumres$MPD~inputs$gd),col=4,lty=2)
#legend("topright",col=3:4,lty=2,c("old","new"),inset=0.05)
#YTR
plot(inputs$gd, sumres$YTR, type="n",col=3, xlab="gestation duration",ylab="YTR")
#plot(c(sumres$peo,sumres$Pexp), c(sumres$YTRo,sumres$YTR), type="n",col=3, xlab="P(exposed)",ylab="YTR")
#points(sumres$peo, sumres$YTRo, col=3)
#abline(lm(sumres$YTRo~sumres$peo),col=3,lty=2)
points(inputs$gd, sumres$YTR,col=4)
abline(lm(sumres$YTR~inputs$gd),col=4,lty=2)
#legend("topright",col=3:4,lty=2,c("old","new"),inset=0.05)
```

and survival reduction

```{r}
#if you uncomment several lines  below you see the plot contrasting with DWH MMIQT (2015) results
#LCY
plot(inputs$sr,sumres$LCY, type="n",col=3, xlab="survival reduction",ylab="LCY")
#plot(c(sumres$peo,sumres$Pexp), c(sumres$LCYo,sumres$LCY), type="n",col=3, xlab="P(exposed)",ylab="LCY")
#points(sumres$peo, sumres$LCYo, col=3)
#abline(lm(sumres$LCYo~sumres$peo),col=3,lty=2)
points(inputs$sr, sumres$LCY,col=4)
abline(lm(sumres$LCY~inputs$sr),col=4,lty=2)
#legend("topright",col=3:4,lty=2,c("old","new"),inset=0.05)
#MPD
plot(inputs$sr, sumres$MPD, type="n",col=3, xlab="survival reduction",ylab="MPD")
#plot(c(sumres$peo,sumres$Pexp), c(sumres$MPDo,sumres$MPD), type="n",col=3, xlab="P(exposed)",ylab="MPD")
#points(sumres$peo, sumres$MPDo, col=3)
#abline(lm(sumres$MPDo~sumres$peo),col=3,lty=2)
points(inputs$sr, sumres$MPD,col=4)
abline(lm(sumres$MPD~inputs$sr),col=4,lty=2)
#legend("topright",col=3:4,lty=2,c("old","new"),inset=0.05)
#YTR
plot(inputs$sr, sumres$YTR, type="n",col=3, xlab="survival reduction",ylab="YTR")
#plot(c(sumres$peo,sumres$Pexp), c(sumres$YTRo,sumres$YTR), type="n",col=3, xlab="P(exposed)",ylab="YTR")
#points(sumres$peo, sumres$YTRo, col=3)
#abline(lm(sumres$YTRo~sumres$peo),col=3,lty=2)
points(inputs$sr, sumres$YTR,col=4)
abline(lm(sumres$YTR~inputs$sr),col=4,lty=2)
#legend("topright",col=3:4,lty=2,c("old","new"),inset=0.05)
```



## Explaining injury metrics across species

### LCY

As noted above, one can anticipate that LCY would be strongly dependent on the initial population size and the proportion exposed. Therefore, we fit a linear model, treating species as sampling units, to explain LCY as a function of these two quantities. The summary of the corresponding linear model follows. Then we present a plot representing the observed values of LCY and those obtained by predicting from fitting a linear model of LCY as explained by initial population size and the proportion exposed.  

```{r}
lmLCY <- lm(LCY ~ Pexp + Nstart, data = sumres)
summary(lmLCY)
plot(sumres$LCY, predict(lmLCY))
```

The LCY is very well explained by these two variable alone. A bit surprisingly, adding the gestation duration (that scales survival across species) does not even seem to help in explaining the injury

```{r}
lmLCY2 <- lm(LCY ~ Pexp + Nstart + inputs$gd, data = sumres)
summary(lmLCY2)
plot(sumres$LCY, predict(lmLCY2))
```

The same seems to happen for the survival reduction, i.e., once population size and proportion exposed are accounted for, the remaining variation in LCY is negligible.

```{r}
lmLCY3 <- lm(LCY ~ Pexp + Nstart + inputs$sr, data = sumres)
summary(lmLCY3)
plot(sumres$LCY, predict(lmLCY3))
```

We note that, when compared to the earlier assessments, the injury seems now considerably less than before. A large part of this was a bug in the calculation in the MMIQT values for LCY, but a part of it might be due to the modeling, and the many changes implemented from then to now.

Using the model considered best, what is the component of variability explained by each variable?

```{r}
anova(lmLCY)
```

The proportion of the variability explained by each input retained in the regression model and the remaining unexplained variation is shown below

```{r}
100*anova(lmLCY)$'Sum Sq'/sum(anova(lmLCY)$'Sum Sq')
```


### YTR

Same idea for YTR, checking what determines that 

```{r}
lmYTR <- lm(YTR ~ Pexp + Nstart + inputs$gd + inputs$sr, data = sumres)
summary(lmYTR)
plot(sumres$YTR, predict(lmYTR))
```

removing initial population size and survival reduction just to check if it helps

```{r}
lmYTR <- lm(YTR ~ Pexp + inputs$gd, data = sumres)
summary(lmYTR)
plot(sumres$YTR, predict(lmYTR))
```


```{r}
anova(lmYTR)
```


The proportion of the variability explained by each input retained in the regression model and the remaining unexplained variation is shown below

```{r}
100*anova(lmYTR)$'Sum Sq'/sum(anova(lmYTR)$'Sum Sq')
```


### MPD

```{r}
lmMPD <- lm(MPD ~ Pexp + Nstart + inputs$gd + inputs$sr, data = sumres)
summary(lmMPD)
plot(sumres$MPD, predict(lmMPD))
```

removing initial population size  just to check if it helps

```{r}
lmMPD <- lm(MPD ~ Pexp + inputs$gd + inputs$sr, data = sumres)
summary(lmMPD)
plot(sumres$MPD, predict(lmMPD))
```


```{r}
anova(lmMPD)
```


The proportion of the variability explained by each input retained in the regression model and the remaining unexplained variation is shown below

```{r}
100*anova(lmMPD)$'Sum Sq'/sum(anova(lmMPD)$'Sum Sq')
```



## Species by species discussion

Here we compare the results for each of the species with those obtained before, trying to shed light on what the differences found might be due to.

To do so we show again the before and after key values for each species.

### Bwsp beaked whales	Beaked whales spp

We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Bwsp"])` to `r round(sumres$LCY[sumres$Sp=="Bwsp"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Bwsp"],3)` to `r round(sumres$MPD[sumres$Sp=="Bwsp"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Bwsp"],1)` to `r round(sumres$YTR[sumres$Sp=="Bwsp"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Bwsp"],3)` to `r round(sumres$Nstart[sumres$Sp=="Bwsp"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Bwsp"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Bwsp"],3)`.

### Fatt pygmy killer whale	*Feresa attenuata*

We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Fatt"])` to `r round(sumres$LCY[sumres$Sp=="Fatt"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Fatt"],3)` to `r round(sumres$MPD[sumres$Sp=="Fatt"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Fatt"],1)` to `r round(sumres$YTR[sumres$Sp=="Fatt"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Fatt"],3)` to `r round(sumres$Nstart[sumres$Sp=="Fatt"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Fatt"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Fatt"],3)`.



### Ggri Risso's dolphin	*Grampus griseus*


We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Ggri"])` to `r round(sumres$LCY[sumres$Sp=="Ggri"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Ggri"],3)` to `r round(sumres$MPD[sumres$Sp=="Ggri"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Ggri"],1)` to `r round(sumres$YTR[sumres$Sp=="Ggri"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Ggri"],3)` to `r round(sumres$Nstart[sumres$Sp=="Ggri"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Ggri"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Ggri"],3)`.


### Gmac short-finned pilot whale	*Globicephala macrorhynchus*


We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Gmac"])` to `r round(sumres$LCY[sumres$Sp=="Gmac"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Gmac"],3)` to `r round(sumres$MPD[sumres$Sp=="Gmac"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Gmac"],1)` to `r round(sumres$YTR[sumres$Sp=="Gmac"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Gmac"],3)` to `r round(sumres$Nstart[sumres$Sp=="Gmac"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Gmac"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Gmac"],3)`.


### Kosp kogia	*Kogia* sp.


We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Kosp"])` to `r round(sumres$LCY[sumres$Sp=="Kosp"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Kosp"],3)` to `r round(sumres$MPD[sumres$Sp=="Kosp"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Kosp"],1)` to `r round(sumres$YTR[sumres$Sp=="Kosp"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Kosp"],3)` to `r round(sumres$Nstart[sumres$Sp=="Kosp"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Kosp"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Kosp"],3)`.


### Pele melon-headed whale	*Peponocephala electra*

We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Pele"])` to `r round(sumres$LCY[sumres$Sp=="Pele"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Pele"],3)` to `r round(sumres$MPD[sumres$Sp=="Pele"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Pele"],1)` to `r round(sumres$YTR[sumres$Sp=="Pele"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Pele"],3)` to `r round(sumres$Nstart[sumres$Sp=="Pele"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Pele"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Pele"],3)`.

### Pmac sperm whale	*Physeter macrocephalus*


We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Pmac"])` to `r round(sumres$LCY[sumres$Sp=="Pmac"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Pmac"],3)` to `r round(sumres$MPD[sumres$Sp=="Pmac"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Pmac"],1)` to `r round(sumres$YTR[sumres$Sp=="Pmac"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Pmac"],3)` to `r round(sumres$Nstart[sumres$Sp=="Pmac"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Pmac"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Pmac"],3)`.

For Pmac the the lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Pmac"],3)` to `r round(sumres$LCY[sumres$Sp=="Pmac"])`, but in fact the true value bug apart was 5637 (cf. GOMOSES 2020 poster). Therefore we would expect a decrease of the LCY by more than 50% just due to the bug (in the specific case of the sperm whale: `r round((1-5637/sumres$LCYo[sumres$Sp=="Pmac"])*100,3)` %).

### Satt pantropical spotted dolphin *Stenella attenuata*

We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Satt"])` to `r round(sumres$LCY[sumres$Sp=="Satt"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Satt"],3)` to `r round(sumres$MPD[sumres$Sp=="Satt"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Satt"],1)` to `r round(sumres$YTR[sumres$Sp=="Satt"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Satt"],3)` to `r round(sumres$Nstart[sumres$Sp=="Satt"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Satt"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Satt"],3)`.

We note that the assumed initial population size increased considerably compared to the MMIQT, which per se would lead to an increase in LCY, everything else being the same (`r round(sumres$LCYo[sumres$No=="Satt"],3)` to `r round(sumres$LCYo[sumres$Nstart=="Satt"],3)`).

While a drop in the LCY can be explained by the already mentioned bug in the LCY calculation, and the reduction in the proportion exposed is also relevant (before it was `r round(sumres$peo[sumres$Sp=="Satt"],3)`, now it is `r round(sumres$Pexp[sumres$Sp=="Satt"],3)`, these do not seem enough to justify the observed differences, namely in the YTR.

### Sbre rough-toothed dolphin	*Steno bredanensis*

We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Sbre"])` to `r round(sumres$LCY[sumres$Sp=="Sbre"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Sbre"],3)` to `r round(sumres$MPD[sumres$Sp=="Sbre"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Sbre"],1)` to `r round(sumres$YTR[sumres$Sp=="Sbre"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Sbre"],3)` to `r round(sumres$Nstart[sumres$Sp=="Sbre"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Sbre"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Sbre"],3)`.

The major difference observed for the proportion exposed could explain most of the observed differences.

### Scly Clymene dolphin	*Stenella clymene*

We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Scly"])` to `r round(sumres$LCY[sumres$Sp=="Scly"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Scly"],3)` to `r round(sumres$MPD[sumres$Sp=="Scly"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Scly"],1)` to `r round(sumres$YTR[sumres$Sp=="Scly"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Scly"],3)` to `r round(sumres$Nstart[sumres$Sp=="Scly"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Scly"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Scly"],3)`.

### Scoe striped dolphin	*Stenella coeruleoalba*

We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Scoe"])` to `r round(sumres$LCY[sumres$Sp=="Scoe"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Scoe"],3)` to `r round(sumres$MPD[sumres$Sp=="Scoe"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Scoe"],1)` to `r round(sumres$YTR[sumres$Sp=="Scoe"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Scoe"],3)` to `r round(sumres$Nstart[sumres$Sp=="Scoe"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Scoe"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Scoe"],3)`.

### Sfro Atlantic spotted dolphin	*Stenella frontalis*

For Sfro the comparison is not direct because in the MMIQT document the Ttrs and the Sfro were combined as "shelf dolphins", and here we considered each stock separately.

We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Sfro"])` to `r round(sumres$LCY[sumres$Sp=="Sfro"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Sfro"],3)` to `r round(sumres$MPD[sumres$Sp=="Sfro"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Sfro"],1)` to `r round(sumres$YTR[sumres$Sp=="Sfro"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Sfro"],3)` to `r round(sumres$Nstart[sumres$Sp=="Sfro"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Sfro"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Sfro"],3)`.

### Slon spinner dolphin	*Stenella longirostris*

We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Slon"])` to `r round(sumres$LCY[sumres$Sp=="Slon"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Slon"],3)` to `r round(sumres$MPD[sumres$Sp=="Slon"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Slon"],1)` to `r round(sumres$YTR[sumres$Sp=="Slon"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Slon"],3)` to `r round(sumres$Nstart[sumres$Sp=="Slon"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Slon"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Slon"],3)`.

### Ttro offshore bottlenose dolphins	*Tursiops truncatus*

We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Ttro"])` to `r round(sumres$LCY[sumres$Sp=="Ttro"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Ttro"],3)` to `r round(sumres$MPD[sumres$Sp=="Ttro"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Ttro"],1)` to `r round(sumres$YTR[sumres$Sp=="Ttro"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Ttro"],3)` to `r round(sumres$Nstart[sumres$Sp=="Ttro"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Ttro"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Ttro"],3)`.

### Ttrs shelf bottlenose dolphins *Tursiops truncatus*

For Ttrs the comparison is not direct because in the MMIQT document the Ttrs and the Sfro were combined as "shelf dolphins", and here we considered each stock separately.

We observe the following changes in injury and in key quantities that might affect the injury (initial population size and the proportion of the population exposed):

* lost cetacean years came down from `r round(sumres$LCYo[sumres$Sp=="Ttrs"])` to `r round(sumres$LCY[sumres$Sp=="Ttrs"])` 
* maximum proportional decrease came down from `r round(sumres$MPDo[sumres$Sp=="Ttrs"],3)` to `r round(sumres$MPD[sumres$Sp=="Ttrs"],3)`
* years to recovery came down from `r round(sumres$YTRo[sumres$Sp=="Ttrs"],1)` to `r round(sumres$YTR[sumres$Sp=="Ttrs"],1)`
* The assumed initial population size changed from `r round(sumres$No[sumres$Sp=="Ttrs"],3)` to `r round(sumres$Nstart[sumres$Sp=="Ttrs"],3)`
* The proportion of the population exposed before was `r round(sumres$peo[sumres$Sp=="Ttrs"],3)` and it is now `r round(sumres$Pexp[sumres$Sp=="Ttrs"],3)`.

# References

DWH MMIQT 2015, Models and analyses for the quantification of injury to Gulf of Mexico cetaceans from the Deepwater Horizon Oil Spill, MM_TR.01_Schwacke_Quantification.of.lnjury.to.GOM.Cetaceans [LINK](https://www.fws.gov/doiddata/dwh-ar-documents/876/DWH-AR0105866.pdf).

Schwacke, L.H., L. Thomas, R.S. Wells, W.E. McFee, A.A. Hahn, K.D. Mullin, E.S. Zolman, B.M. Quigley, T.K. Rowles and J.H. Schwacke. 2017. An age-, sex- and class-structured population model for estimating nearshore common bottlenose dolphin injury following the Deepwater Horizon oil spill. *Endangered Species Research* **33**: 265-279. DOI: [10.3354/esr00777](https://doi.org/10.3354/esr00777).