---
title: Results accompanying manuscript "Population model predicts catastrophic impact of proposed habitat restoration project for common bottlenose dolphin (Tursiops truncatus) in Barataria Bay, Louisiana, USA"
author: Len Thomas$^1$, Tiago Marques$^1$, Cormac Booth$^2$, Ryan Takeshita$^3$ and
  Lori Schwacke$^3$
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, fig.width = 7, fig.height = 5, fig.align = 'center')
```

\begin{center}
$^1$Centre for Research into Ecological and Environmental Modelling, University of St Andrews. 
$^2$SMRU Consulting. $^3$National Marine Mammal Foundation. 
\end{center}

```{r}
#Define some constants
alpha <- 0.05
diversion.start.year <- 2027 - 2010 + 1
diversion.n.years <- 50
diversion.end.year <- (2027 + diversion.n.years - 1) - 2010 + 1
functional.extinction <- 30
yrs.after.diversion <- 10

#Read in population run results -- creates objects res.mort and res.pop
load(file = "out.Rdata")
#Truncate to just the part we are interested in
res.mort <- res.mort[1:diversion.end.year, , , ]
res.pop <- res.pop[1:diversion.end.year, , , ]

year <- 2010 : (2010 - 1 + dim(res.mort)[1])

#Read in stratum survival change
strata <- c("Island", "Southeast", "Central", "West")
n.strata <- length(strata)
LowSalS <- read.csv("BB_AllRegions_Survival.csv", header = TRUE)
# Percentage survival difference per stratum
PSD <- ((LowSalS[LowSalS$scenario == "APA", strata] - LowSalS[LowSalS$scenario == "NAA", strata]) / LowSalS[LowSalS$scenario == "NAA", strata]) * 100
```

```{r}
#Compute appropriate summaries

#Population summaries
pop.strat.med <- apply(res.pop, c(1, 3, 4), median)
pop.strat.lcl <- apply(res.pop, c(1, 3, 4), quantile, alpha / 2)
pop.strat.ucl <- apply(res.pop, c(1, 3, 4), quantile, 1 - alpha / 2)
pop.tot <- apply(res.pop, c(1, 2, 4), sum)
pop.tot.med <- apply(pop.tot, c(1, 3), median)
pop.tot.lcl <- apply(pop.tot, c(1, 3), quantile, alpha / 2)
pop.tot.ucl <- apply(pop.tot, c(1, 3), quantile, 1 - alpha / 2)

#Mortality summary - just need year of diversion
excess.mort.diversion.start <- res.mort[diversion.start.year, , , 1] - 
  res.mort[diversion.start.year, , , 2]
excess.mort.strat.diversion.start.med <- apply(excess.mort.diversion.start, 2, median)
excess.mort.strat.diversion.start.lcl <- apply(excess.mort.diversion.start, 2, quantile, alpha / 2)
excess.mort.strat.diversion.start.ucl <- apply(excess.mort.diversion.start, 2, quantile, 1 - alpha / 2)
excess.mort.diversion.start <- apply(excess.mort.diversion.start, 1, sum)
excess.mort.tot.diversion.start.med <- median(excess.mort.diversion.start)
excess.mort.tot.diversion.start.lcl <- quantile(excess.mort.diversion.start, alpha / 2)
excess.mort.tot.diversion.start.ucl <- quantile(excess.mort.diversion.start, 1 - alpha / 2)
```

# Summary of findings

1. The proposed Mid-Barataria Sediment Diversion (MBSD) project will result in decreased levels of salinity in Barataria Bay, Louisiana. This decreased salinity has been predicted by the National Oceanographic and Atmospheric Administration (NOAA) to cause increased mortality of bottlenose dolphins in the Barataria Bay Estuarine System (BBES) Stock.  

2. We used an existing model for the population dynamics of this stock to predict the population consequences of the increased mortality.  We compared population projections under two scenarios (described in the Draft Environmental Impact Statement for the proposed project), "Applicants Preferred Alternative" (APA) and "No Action Alternative" (NAA), using the same four geographic regions ("strata") as NOAA and assuming no movement of animals among strata.

```{r}
#Calculate population decline statistics used in next bullet
pop.ch.apa <- ((pop.tot[diversion.start.year + 1, , 1] -
                 pop.tot[diversion.start.year, , 1]) /
                 pop.tot[diversion.start.year, , 1])
pop.ch.apa.med <- median(pop.ch.apa)
pop.ch.apa.lcl <- quantile(pop.ch.apa, alpha / 2)
pop.ch.apa.ucl <- quantile(pop.ch.apa, 1 - alpha / 2)
pop.ch.naa <- ((pop.tot[diversion.start.year + 1, , 2] -
                 pop.tot[diversion.start.year, , 2]) /
                 pop.tot[diversion.start.year, , 2])
pop.ch.naa.med <- median(pop.ch.naa)
pop.ch.naa.lcl <- quantile(pop.ch.naa, alpha / 2)
pop.ch.naa.ucl <- quantile(pop.ch.naa, 1 - alpha / 2)
pop.apa.vs.naa <- (pop.tot[diversion.start.year + 1, , 1] /
                        pop.tot[diversion.start.year + 1, , 2])
pop.apa.vs.naa.med <- median(pop.apa.vs.naa)
pop.apa.vs.naa.lcl <- quantile(pop.apa.vs.naa, alpha / 2)
pop.apa.vs.naa.ucl <- quantile(pop.apa.vs.naa, 1 - alpha / 2)
```

3. The model predicts an immediate and severe population-level decline under the APA.  In the first year of operation under the APA (2027), median predicted excess mortality under the APA is `r round(excess.mort.tot.diversion.start.med)` dolphins (95% confidence interval [CI] `r round(excess.mort.tot.diversion.start.lcl)`-`r round(excess.mort.tot.diversion.start.ucl)`), leading to a median stock decline of `r round(-pop.ch.apa.med * 100)`% (95% CI `r round(-pop.ch.apa.ucl * 100)`-`r round(-pop.ch.apa.lcl * 100)`). By contrast, under the NAA the stock is predicted to increase by `r round(pop.ch.naa.med * 100)`% (95% CI `r round(pop.ch.naa.lcl * 100)`-`r round(pop.ch.naa.ucl * 100)`) -- the increase is because the stock is estimated to still be in recovery from the *Deepwater Horizon* oil spill.  Therefore, after one year of operation, the stock is predicted to be `r round((1 - pop.apa.vs.naa.med) * 100)`% smaller (95% CI `r round((1 - pop.apa.vs.naa.ucl) * 100)`-`r round((1 - pop.apa.vs.naa.lcl) * 100)`) under the APA than under the NAA.

```{r}
#Calculate statistics required for next bullet
is.extinct <- function(x, limit){
  return(sum(x < limit) / length(x) )
}
p.extinct.10 <- apply(res.pop[diversion.start.year + yrs.after.diversion, , , 1], 2, is.extinct, limit = functional.extinction)

pop.apa.vs.naa.strat.10 <- res.pop[diversion.start.year + yrs.after.diversion, , , 1] / res.pop[diversion.start.year + yrs.after.diversion, , , 2]
pop.apa.vs.naa.strat.10.med <- apply(pop.apa.vs.naa.strat.10, 2, median)
pop.apa.vs.naa.strat.10.lcl <- apply(pop.apa.vs.naa.strat.10, 2, quantile, alpha / 2)
pop.apa.vs.naa.strat.10.ucl <- apply(pop.apa.vs.naa.strat.10, 2, quantile, 1 - alpha / 2)
```

4. After `r yrs.after.diversion` years of operation, the parts of the stock in the Central and West strata are predicted to be functionally extinct (probability of < `r functional.extinction` animals remaining is `r round(p.extinct.10["Central"], 2)` in the Central stratum and `r round(p.extinct.10["West"], 2)` in the West stratum). The part in the Southeast stratum, while not extinct, is predicted to be `r round((1 - pop.apa.vs.naa.strat.10.med["Southeast"]) * 100)`% lower (95% CI `r round((1 - pop.apa.vs.naa.strat.10.ucl["Southeast"]) * 100)`-`r round((1 - pop.apa.vs.naa.strat.10.lcl["Southeast"]) * 100)`) under the APA than under the NAA. The Island stratum is less severely affected with a median predicted decline of `r round((1 - pop.apa.vs.naa.strat.10.med["Island"]) * 100)`% (95% CI `r round((1 - pop.apa.vs.naa.strat.10.ucl["Island"]) * 100)`-`r round((1 - pop.apa.vs.naa.strat.10.lcl["Island"]) * 100)`).

```{r}
#Calculate population decline statistics used in next bullet
pop.apa.vs.naa.strat.end <- res.pop[diversion.end.year, , , 1] / res.pop[diversion.end.year, , , 2]
pop.apa.vs.naa.strat.end.med <- apply(pop.apa.vs.naa.strat.end, 2, median)
pop.apa.vs.naa.strat.end.lcl <- apply(pop.apa.vs.naa.strat.end, 2, quantile, alpha / 2)
pop.apa.vs.naa.strat.end.ucl <- apply(pop.apa.vs.naa.strat.end, 2, quantile, 1 - alpha / 2)

pop.tot.apa.vs.naa.perc <- (1 - pop.tot[diversion.end.year, , 1] / pop.tot[diversion.end.year, , 2]) * 100
```

5. In the 50th (i.e., final) year of operation, dolphins in three out of the four strata are predicted to be functionally extinct under the APA, with the remaining Island stratum being severely reduced relative to the NAA (median predicted population size of Island stratum is `r round((1 - pop.apa.vs.naa.strat.end.med["Island"]) * 100)`% lower [95% CI `r round((1 - pop.apa.vs.naa.strat.end.ucl["Island"]) * 100)`-`r round((1 - pop.apa.vs.naa.strat.end.lcl["Island"]) * 100)`] under the APA than under the NAA).  Overall, by the year `r 2010 + diversion.end.year - 1`, the median predicted stock size across all of Barataria Bay under the APA is `r round(median(pop.tot[diversion.end.year, , 1]))` dolphins (95% CI `r round(quantile(pop.tot[diversion.end.year, , 1], alpha / 2))`-`r round(quantile(pop.tot[diversion.end.year, , 1], 1 - alpha / 2))`) compared to `r round(median(pop.tot[diversion.end.year, , 2]))` (95% CI `r round(quantile(pop.tot[diversion.end.year, , 2], alpha / 2))`-`r round(quantile(pop.tot[diversion.end.year, , 2], 1 - alpha / 2))`) under the NAA.  In other words, the stock is predicted to be `r round(median(pop.tot.apa.vs.naa.perc))`% smaller (95% CI `r round(quantile(pop.tot.apa.vs.naa.perc, alpha / 2))`-`r round(quantile(pop.tot.apa.vs.naa.perc, 1 - alpha / 2))`) under the APA than then NAA.

# Impact on survival from NOAA analysis

We obtained from NOAA 1,000 replicate predictions of estimated annual survival under APA and NAA scenarios in each of the four strata, derived from the model of Garrison et al. (2020).  The replicate predictions represent the range of scientific uncertainty on possible impacts, accounting for factors such as uncertainty on the salinity field for a given set of hydrographic conditions, uncertainty on animal movement and hence exposure, and uncertainty on the effect of low salinity on dolphin survival (see Garrison et al. 2020 for details).  Note that all predictions are based on a single assumed annual hydrograph, that for 1970 (Garrison et al. 2020), and so do not account for uncertainty in future hydrographic conditions (see Discussion in main paper).

For each replicate prediction and stratum, we calculated the percentage difference in survival between the APA and the NAA as follows: 
$$
\mbox{\% difference in survival} = \frac{\mbox{survival under APA - survival under NAA}}{\mbox{survival under NAA}} \times 100
$$

The resulting distribution of percentage difference in survival in each stratum is shown in Figure 1, with associated summary statistics in Table 1. For the Island stratum, the median prediction is of a 2% decline in survival under the APA relative to the NAA, although in 10% of replicates the predicted survival decline is greater than 20%. For the Southeast stratum, the median prediction is of a 14% decline in survival with 40% of replicates predicting a survival decline of greater than 20%. Note, however, that 24% of replicates in this stratum predict an increase in survival under APA relative to the NAA. For the Central and West strata there is a large predicted decline in survival under almost all replicates.

```{r, fig.cap = "Predicted percentage difference in dolphin annual survival under the Applicant Preferred Alternative (APA) compared with the No Action Alternative (NAA). One thousand predicted survival rates were provided by NOAA and were derived from the model of Garrison et al. (2020)."}
par(mfrow = c(2, 2))
for (i in 1:n.strata){
  hist(PSD[, strata[i]], main = strata[i], xlab = "% Difference in Survival", probability = FALSE, xlim = range(PSD))
}
par(mfrow = c(1, 1))
```

```{r}
tmp <- data.frame(stratum = strata)
tmp$median_PSD <- apply(PSD, 2, median)
tmp$perc_lt_0.8 <- apply(PSD, 2, function(x){sum(x < -20) / length(x) * 100})
tmp$perc_gt_1 <- apply(PSD, 2, function(x){sum(x > 0)/length(x) * 100})
kable(tmp, digits = 0, longtable = FALSE, col.names = c("Stratum", "median % diff", "% (diff < -20%)", "% (diff > 0)"), caption = "Summary statistics on predicted percentage difference in dolphin annual survival under the APA relative to the NAA.  First column is median predicted percentage difference, second is percentage of replicates that predict a decline in survival of 20% or more, third is percentage of replicates that predict an increase in survival.")
```

# Population consequences

```{r}
alpha <- 0.05
n.toshow <- 10
```

Figure 2 summarizes the population trajectories over all 1,000 realizations under APA (red) and NAA (black) scenarios.  The populations follow the same trajectory under both scenarios up until 2027, when proposed MBSD operations start.  During this period (2010-2027) the populations experience the negative effect of the DWH oil spill and, starting around 2020, begin to recover. After 2027 under the NAA, the populations continue to recover and reach a steady state long before the end of the simulation time period.  Under the APA the median prediction for the Island stratum is of a steady decline, while the other strata experience rapid declines to extinction.  The prediction at stock level, i.e., summing across strata, is shown in Figure 3.  Under the APA, the stock is predicted to decline precipitously at first and then more gradually, reaching very low levels relative to the NAA by the end of the simulation time period.

```{r, fig.cap = "Summary of predicted population trajectories by stratum under the Applicant Preferred Alternative (red) and No Action Alternative (black) scenarios.  Solid line shows median; dashed lines show 95\\% confidence limits."}
par(mfrow = c(2, 2))
for(i in 1:n.strata){
  ylim <- range(c(pop.strat.med[, i, ], pop.strat.lcl[, i, ], pop.strat.ucl[, i, ]))
  plot(year, pop.strat.med[, i, 2], main = strata[i], type = "l", ylim = ylim, ylab = "dolphins")
  lines(year, pop.strat.lcl[, i, 2], lty = 2)
  lines(year, pop.strat.ucl[, i, 2], lty = 2)
  lines(year, pop.strat.med[, i, 1], col = "red")
  lines(year, pop.strat.lcl[, i, 1], col = "red", lty = 2)
  lines(year, pop.strat.ucl[, i, 1], col = "red", lty = 2)
}
par(mfrow = c(1, 1))
```


```{r, fig.cap = "Summary of predicted stock trajectory under the Applicant Preferred Alternative (red) and No Action Alternative (black) scenarios.  Solid line shows median; dashed lines show 95\\% confidence limits."}
ylim <- range(c(pop.tot.med, pop.tot.lcl, pop.tot.ucl))
plot(year, pop.tot.med[, 2], main = "Barataria Bay Estuarine System Stock", type = "l", ylim = ylim, ylab = "dolphins")
lines(year, pop.tot.lcl[, 2], lty = 2)
lines(year, pop.tot.ucl[, 2], lty = 2)
lines(year, pop.tot.med[, 1], col = "red")
lines(year, pop.tot.lcl[, 1], col = "red", lty = 2)
lines(year, pop.tot.ucl[, 1], col = "red", lty = 2)
```

The summaries given in Figures 2 and 3 are computed from 1,000 random realizations of the model.  Figure 4 shows `r n.toshow` example realizations.  The part of the stock in the Island stratum experiences occasional large population decreases associated with years where there is a large decline in survival under the APA; in most years, however, there is little or no decline.  After 50 years of operation, all realizations have experienced an overall decline and none are at the level of the corresponding NAA.  The part of the stock in the Southeast stratum experiences frequent stronger declines, but also occasional increases associated with survival increase under the APA.  Nevertheless, after 50 years of operation, all realizations are at or close to zero.  The parts of the stock in the Central and West strata experience rapid declines towards zero in all realizations.

```{r, fig.cap = "Ten example realizations of the population simulation under the Applicant Preferred Alternative (red) and No Action Alternative (black) scenarios."}
par(mfrow = c(2, 2))
for(i in 1:n.strata){
  ylim <- range(res.pop[, 1:n.toshow, i, ])
  plot(year, res.pop[, 1, i, 2], main = strata[i], type = "l", ylim = ylim, ylab = "dolphins")
  for(j in 1:n.toshow){
    lines(year, res.pop[, j, i, 2])
    lines(year, res.pop[, j, i, 1], col = "red")
  }
}
par(mfrow = c(1, 1))
```


```{r}
#Work out proportion of population killed by APA
prop.excess <- excess.mort.diversion.start / pop.tot[diversion.start.year, , 1]
perc.excess <- prop.excess * 100
```

Quantitative summaries of the results are given in Tables 2-6^[Note that in all these tables, the median shown in the "Total" row is calculated by first aggregating the strata and then calculating the median.  This is not the same as simply summing the stratum medians.  The same is true for the confidence limits.].  Table 2 shows the predicted mortalities in the first year of the proposed MBSD operation (2027) under APA, NAA and the difference between the two, which represents the predicted excess mortalities under the APA scenario.  In this first year of MBSD operation, the median predicted excess mortality under the APA is `r round(excess.mort.tot.diversion.start.med)` dolphins with 95% CI `r round(excess.mort.tot.diversion.start.lcl)`-`r round(excess.mort.tot.diversion.start.ucl)`.  This excess mortality represents a median of `r format(median(perc.excess), digits = 0)`% of the stock (95% CI `r format(quantile(perc.excess, alpha / 2), digits = 0)`-`r format(quantile(perc.excess, 1 - alpha / 2), digits = 0)`) killed by the MBSD in its first year of operation.

```{r}
#Table 2
make.outstring <- function(est, lcl, ucl){
  return(paste0(est, " (", lcl, " — ", ucl, ")"))
}

res.mort.tot.diversion.start <- apply(res.mort[diversion.start.year, , , ], c(1, 3), sum)

res.mort.apa.med <- round(apply(res.mort[diversion.start.year, , , 1], 2, median))
res.mort.apa.med <- c(res.mort.apa.med, round(median(res.mort.tot.diversion.start[, 1])))
res.mort.apa.lcl <- round(apply(res.mort[diversion.start.year, , , 1], 2, quantile, alpha / 2))
res.more.apa.lcl <- c(res.mort.apa.lcl, round(quantile(res.mort.tot.diversion.start[, 1]), alpha / 2))
res.mort.apa.ucl <- round(apply(res.mort[diversion.start.year, , , 1], 2, quantile, 1 - alpha / 2))
res.more.apa.ucl <- c(res.mort.apa.ucl, round(quantile(res.mort.tot.diversion.start[, 1]), 1 - alpha / 2))
res.mort.naa.med <- round(apply(res.mort[diversion.start.year, , , 2], 2, median))
res.mort.naa.med <- c(res.mort.naa.med, round(median(res.mort.tot.diversion.start[, 2])))
res.mort.naa.lcl <- round(apply(res.mort[diversion.start.year, , , 2], 2, quantile, alpha / 2))
res.more.naa.lcl <- c(res.mort.naa.lcl, round(quantile(res.mort.tot.diversion.start[, 2]), alpha / 2))
res.mort.naa.ucl <- round(apply(res.mort[diversion.start.year, , , 2], 2, quantile, 1 - alpha / 2))
res.more.naa.ucl <- c(res.mort.naa.ucl, round(quantile(res.mort.tot.diversion.start[, 2]), 1 - alpha / 2))
tmp <- data.frame(stratum = c(strata, "Total"))
tmp$apa <- make.outstring(res.mort.apa.med, res.mort.apa.lcl, res.mort.apa.ucl)
tmp$naa <- make.outstring(res.mort.naa.med, res.mort.naa.lcl, res.mort.naa.ucl)
tmp$excess <- make.outstring(round(c(excess.mort.strat.diversion.start.med, excess.mort.tot.diversion.start.med)), round(c(excess.mort.strat.diversion.start.lcl, excess.mort.tot.diversion.start.lcl)), round(c(excess.mort.strat.diversion.start.ucl, excess.mort.tot.diversion.start.ucl)))
kable(tmp, col.names = c("Stratum", "APA mortality", "NAA mortality", "excess mortality (APA-NAA)"), caption = "Predicted dolphin mortality in 2027 under APA and NAA scenarios.  Last column shows excess mortality -- i.e., mortality under APA minus mortality under NAA.  Values are medians from the model simulations, with 95% confidence intervals in brackets.")
```

Estimated stock size in the 2027, before operation of the proposed MBSD, was `r round(pop.tot.med[diversion.start.year, 1])` animals (95% CI `r round(pop.tot.lcl[diversion.start.year, 1])`-`r round(pop.tot.ucl[diversion.start.year, 1])`).  Estimated stock sizes in 2028 under the APA and NAA are shown in Table 3.  Under the APA, the stock is predicted to decline by `r round(-pop.ch.apa.med * 100)`% (95% CI `r round(-pop.ch.apa.ucl * 100)`-`r round(-pop.ch.apa.lcl * 100)`) due to mortalities caused by the MBSD operation.  By contrast, under the NAA the stock is predicted to increase by  `r round(pop.ch.naa.med * 100)`% (95% CI `r round(pop.ch.naa.lcl * 100)`-`r round(pop.ch.naa.ucl * 100)`) -- the increase is because the stock is estimated be still in recovery from the DWH oil spill.  Therefore, by the end of the first year of MBSD operations, the stock is predicted to be `r round((1 - pop.apa.vs.naa.med) * 100)`% smaller (95% CI `r round((1 - pop.apa.vs.naa.ucl) * 100)`-`r round((1 - pop.apa.vs.naa.lcl) * 100)`) under the APA than under the NAA (Table 3).


```{r}
get.str.pop.size <- function(res.pop, strata, stratum, yr, alpha){
  #Returns median population size under diversion and no_diversion, and perc.diff
  # in yr y; also alpha-level CI. All rounded to nearest integer and formatted
  # for inclusion in a table.
  #If stratum = 0 then for all strata, otherwise for particular stratum
  if(stratum == 0) {
    res.pop <- apply(res.pop[yr, , , ], c(1, 3), sum)
    str <- "Total"
  } else {
    res.pop <- res.pop[yr, , stratum, ]
    str <- strata[stratum]
  }
  perc.diff <- (res.pop[, 1] - res.pop[, 2]) / res.pop[, 2] * 100
  return(list(stratum = str, 
         apa = paste0(round(median(res.pop[, 1])), " (",
                      round(quantile(res.pop[, 1], alpha / 2)), " — ",
                      round(quantile(res.pop[, 1], 1 - alpha / 2)), ")"),  
         naa = paste0(round(median(res.pop[, 2])), " (",
                      round(quantile(res.pop[, 2], alpha / 2)), " — ",
                      round(quantile(res.pop[, 2], 1 - alpha / 2)), ")"),  
         perc.diff = paste0(round(median(perc.diff)), " (",
                            round(quantile(perc.diff, alpha / 2))," — ",
                            round(quantile(perc.diff, 1 - alpha / 2)), ")")))
}
get.pop.size <- function(res.pop, yr, strata, alpha){
  #Wrapper function to get pop size for a particular year
  tmp <- data.frame(get.str.pop.size(res.pop, strata, 1, yr, alpha))
  for(i in 2:n.strata){
    tmp <- rbind(tmp, get.str.pop.size(res.pop, strata, i, yr, alpha))
  }
  tmp <- rbind(tmp, get.str.pop.size(res.pop, strata, 0, yr, alpha))
  return(tmp)
}
```
```{r}
#Table 3
tmp <- get.pop.size(res.pop, diversion.start.year + 1, strata, alpha)
kable(tmp, col.names = c("Stratum", "APA dolphins", "NAA dolphins", "% difference"), caption = "Predicted number of dolphins in 2028 (after 1 year of operation of the MBSD) by stratum and overall under APA and NAA scenarios, and percentage difference between scenarios.  Values are medians with 95% confidence intervals in brackets.")
```

Tables 4 and 5 show the predicted population size in 2038 and 2076 respectively (i.e., after 10 years of operation of the MBSD and in the last year of the 50 year planning horizon) under APA and NAA, as well as the difference between the two scenarios.

\newpage

```{r}
#Table 4
tmp <- get.pop.size(res.pop, diversion.start.year + yrs.after.diversion, strata, alpha)
kable(tmp, col.names = c("Stratum", "APA dolphins", "NAA dolphins", "% difference"), caption = "Predicted number of dolphins in 2038 (after 10 years of operation of the MBSD) by stratum and overall under APA and NAA scenarios, and percentage difference between scenarios.  Values are medians with 95% confidence intervals in brackets.")
```

```{r}
#Table 5
tmp <- get.pop.size(res.pop, diversion.end.year, strata, alpha)
kable(tmp, col.names = c("Stratum", "APA dolphins", "NAA dolphins", "% difference"),  caption = "Predicted number of dolphins in 2076 (at the end of the planning horizon for the MBSD) by stratum and overall under APA and NAA scenarios, and percentage difference between scenarios.  Values are medians with 95% confidence intervals in brackets.")
```

Table 6 shows the predicted probability of functional extinction (i.e., proportion of simulation runs where the number of dolphins is less than `r functional.extinction`) in each stratum in 2038 and 2076.

```{r}
p.extinct.end <- apply(res.pop[diversion.end.year, , , 1], 2, is.extinct, limit = functional.extinction)
tmp <- data.frame(stratum = strata, p.10 = p.extinct.10, p.end = p.extinct.end)
rownames(tmp) <- NULL
kable(tmp, row.names = NA, col.names = c("Stratum", "p(extinct) in 2038", "p(extinct) in 2076"), digits = 2, caption = "Predicted probability of functional extinction (i.e., fewer than 30 dolphins remaining) by stratum in 2038 (after 10 years of operation of the MBSD under APA) and 2076 (at the end of the planning horizon for the MBSD)")
```

# Literature cited

Garrison, L.P., J. Litz and C. Sinclair. 2020. Predicting the effects of low salinity associated with the MBSD project on resident common bottlenose dolphins (_Tursiops truncatus_) in Barataria Bay, LA. NOAA Technical Memorandum NMFS-SEFSC-748

